---
title: "Data Loader - Paralog Chromatin Accessibility Analysis"
author: "PARE Pipeline"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Overview

This script loads and prepares the primary dataframes for the paralog chromatin accessibility analysis.

**Output dataframes saved to RData:**
- `re_table` - REs with cluster assignments, FIRE ratios, per-RE allelic diff, and group-level allelic stats
- `re_pairs` - ALL pairwise RE combinations within groups, with alignment info and group-level paralog stats
- `group_stats` - One row per group combining paralog stats (from re_pairs) and allelic stats (from re_table)

## Load Libraries

```{r libraries}
library(tidyverse)
library(ggplot2)
library(glue)
library(data.table)
library(hexbin)
library(scales)

# Install mrvplot if not available
if (!requireNamespace("mrvplot", quietly = TRUE)) {
    message("Installing mrvplot from GitHub...")
    devtools::install_github("mrvollger/mrvplot")
}
library(mrvplot)

# Use mrvplot theme
theme_set(mrv_grid())

# Color palette for paralog vs allelic comparisons
COLOR_PARALOG <- "darkred"
COLOR_ALLELIC <- "gray40"

# Create output directory for figures
FIGURE_DIR <- "figures"
if (!dir.exists(FIGURE_DIR)) {
    dir.create(FIGURE_DIR, recursive = TRUE)
}

# Create output directory for data
DATA_DIR <- "data"
if (!dir.exists(DATA_DIR)) {
    dir.create(DATA_DIR, recursive = TRUE)
}

# Minimum coverage threshold
min_coverage <- 10
```

## Load RE Table

Load the RE table with cluster assignments and peak/FIRE information,
then calculate FIRE ratios and filter for sufficient coverage.

```{r load-re-table}
# Load REs with cluster assignments and peak information
re_table_raw <- read_tsv(
    "results/pileup/peaks_with_clusters.bed.gz",
    col_types = cols(.default = col_guess())
) %>%
    rename(chrom = `#chrom`) %>%
    mutate(
        # Calculate FIRE ratio (proportion of fibers with accessible chromatin)
        fire_ratio_ref = fire_coverage_ref / coverage_ref,
        fire_ratio_alt = fire_coverage_alt / coverage_alt,
        # Handle division by zero
        fire_ratio_ref = if_else(coverage_ref == 0, NA_real_, fire_ratio_ref),
        fire_ratio_alt = if_else(coverage_alt == 0, NA_real_, fire_ratio_alt)
    )

cat("Loaded", nrow(re_table_raw), "REs\n")
cat("Unique clusters:", n_distinct(re_table_raw$cluster_id), "\n")
cat("Samples:", paste(unique(re_table_raw$sample), collapse = ", "), "\n")

# Filter for sufficient coverage
re_table_coverage_filtered <- re_table_raw %>%
    filter(coverage_ref >= min_coverage)

cat("REs with coverage >=", min_coverage, ":", nrow(re_table_coverage_filtered), "\n")
cat("REs filtered out:", nrow(re_table_raw) - nrow(re_table_coverage_filtered), "\n")

# Filter to keep only groups (cluster, sample, haplotype) with at least 2 REs
# This is essential for paralog analysis - we need at least 2 REs to compare
groups_with_multiple_res <- re_table_coverage_filtered %>%
    count(cluster_id, sample, haplotype) %>%
    filter(n >= 2) %>%
    select(-n)

re_table <- re_table_coverage_filtered %>%
    semi_join(groups_with_multiple_res, by = c("cluster_id", "sample", "haplotype")) %>%
    # Calculate per-RE allelic difference (ref vs alt at this position)
    mutate(
        allelic_diff = abs(fire_ratio_ref - fire_ratio_alt),
        has_allelic_data = !is.na(fire_ratio_ref) & !is.na(fire_ratio_alt) & coverage_alt >= min_coverage
    ) %>%
    # Add group-level allelic statistics using grouped mutate
    group_by(cluster_id, sample, haplotype) %>%
    mutate(
        # Count of REs in this group
        n_paralogs = n(),
        # Allelic stats (only for REs with sufficient alt coverage)
        n_res_allelic = sum(has_allelic_data, na.rm = TRUE),
        group_mean_allelic_diff = mean(allelic_diff[has_allelic_data], na.rm = TRUE),
        group_sd_allelic_diff = sd(allelic_diff[has_allelic_data], na.rm = TRUE),
        group_cv_allelic_diff = group_sd_allelic_diff / group_mean_allelic_diff
    ) %>%
    ungroup()

n_singleton_groups <- nrow(re_table_coverage_filtered) - nrow(re_table)
cat("REs in singleton groups removed:", n_singleton_groups, "\n")
cat("REs after group filter:", nrow(re_table), "\n")
cat("REs with allelic data (sufficient alt coverage):", sum(re_table$has_allelic_data), "\n")

glimpse(re_table)
```

## Generate All Pairwise Paralog Combinations

Generate ALL pairwise combinations of REs within the same (cluster, sample, haplotype) group.
This ensures we have every possible paralog pair, not just those with alignments.
Alignment info (like sequence identity) is joined when available.

```{r re-pairs}
# Load alignment data for joining
alignment_data <- read_tsv(
    "results/all_by_all/re_pairwise_identity.tsv.gz",
    col_types = cols(
        query_re_id = col_character(),
        target_re_id = col_character(),
        query_cluster = col_character(),
        target_cluster = col_character(),
        same_cluster = col_logical(),
        query_sample = col_character(),
        target_sample = col_character(),
        query_haplotype = col_character(),
        target_haplotype = col_character(),
        .default = col_guess()
    )
) %>%
    # Filter for paralog pairs: same cluster, same sample, same haplotype
    filter(
        same_cluster == TRUE,
        query_sample == target_sample,
        query_haplotype == target_haplotype
    ) %>%
    select(query_re_id, target_re_id, gap_compressed_identity, alignment_length, num_matches)

cat("Alignment records loaded:", nrow(alignment_data), "\n")

# Create alignment lookup with both directions
alignment_info <- alignment_data %>%
    bind_rows(
        alignment_data %>%
            select(
                query_re_id = target_re_id,
                target_re_id = query_re_id,
                gap_compressed_identity, alignment_length,
                num_matches
            )
    )

# Generate ALL pairwise paralog combinations from re_table
# This uses a self-join to create every possible pair
re_pairs <- re_table %>%
    select(cluster_id, sample, haplotype, re_id, fire_ratio_ref, fire_ratio_alt) %>%
    rename(
        query_re_id = re_id,
        fire_ratio_ref_query = fire_ratio_ref,
        fire_ratio_alt_query = fire_ratio_alt
    ) %>%
    inner_join(
        re_table %>%
            select(cluster_id, sample, haplotype,
                target_re_id = re_id,
                fire_ratio_ref_target = fire_ratio_ref,
                fire_ratio_alt_target = fire_ratio_alt
            ),
        by = c("cluster_id", "sample", "haplotype"),
        relationship = "many-to-many"
    ) %>%
    # Keep only pairs where query < target to avoid duplicates and self-pairs
    filter(query_re_id < target_re_id) %>%
    # Calculate per-pair metrics
    mutate(
        fire_diff = abs(fire_ratio_ref_query - fire_ratio_ref_target),
        # Allelic differences for query and target REs
        allelic_diff_query = abs(fire_ratio_ref_query - fire_ratio_alt_query),
        allelic_diff_target = abs(fire_ratio_ref_target - fire_ratio_alt_target)
    ) %>%
    # Join alignment info when available
    left_join(
        alignment_info,
        by = c("query_re_id", "target_re_id")
    ) %>%
    # Add identity bin (pre-computed for downstream analyses)
    mutate(
        identity_bin = cut(
            gap_compressed_identity,
            breaks = c(0, 0.90, 0.95, 0.98, 0.99, 0.999, 1.0),
            labels = c("<90%", "90-95%", "95-98%", "98-99%", "99-99.9%", ">99.9%"),
            include.lowest = TRUE
        )
    ) %>%
    # Add group-level statistics using grouped mutate
    group_by(cluster_id, sample, haplotype) %>%
    mutate(
        # Number of pairs in this group (n_paralogs = pairs + 1 for groups with >=2 REs)
        n_pairs = n(),
        # Note: n_paralogs (actual RE count) requires counting unique REs
        n_paralogs = n_distinct(c(query_re_id, target_re_id)),
        # Group-level FIRE difference stats
        group_mean_fire_diff = mean(fire_diff, na.rm = TRUE),
        group_sd_fire_diff = sd(fire_diff, na.rm = TRUE),
        group_median_fire_diff = median(fire_diff, na.rm = TRUE),
        group_min_fire_diff = min(fire_diff, na.rm = TRUE),
        group_max_fire_diff = max(fire_diff, na.rm = TRUE),
        group_range_fire_diff = group_max_fire_diff - group_min_fire_diff,
        # Group-level identity stats (from alignments)
        group_mean_identity = mean(gap_compressed_identity, na.rm = TRUE),
        group_median_identity = median(gap_compressed_identity, na.rm = TRUE),
        group_min_identity = min(gap_compressed_identity, na.rm = TRUE),
        group_max_identity = max(gap_compressed_identity, na.rm = TRUE),
        # Group-level allelic difference stats
        group_mean_allelic_diff = mean(c(allelic_diff_query, allelic_diff_target), na.rm = TRUE),
        group_sd_allelic_diff = sd(c(allelic_diff_query, allelic_diff_target), na.rm = TRUE)
    ) %>%
    ungroup()

cat("\n=== Paralog Pairs (all pairwise combinations) ===\n")
cat("Total pairs:", nrow(re_pairs), "\n")
cat("Pairs with alignment data:", sum(!is.na(re_pairs$gap_compressed_identity)), "\n")
cat("Pairs without alignment:", sum(is.na(re_pairs$gap_compressed_identity)), "\n")
cat("Unique groups:", n_distinct(re_pairs$cluster_id, re_pairs$sample, re_pairs$haplotype), "\n")

cat("\n=== Pre-computed columns added ===\n")
cat("Per-pair: fire_diff, allelic_diff_query, allelic_diff_target, identity_bin\n")
cat("Per-group: n_pairs, n_paralogs, group_mean/sd/median/min/max_fire_diff,\n")
cat("           group_mean/median/min/max_identity, group_mean/sd_allelic_diff\n")

glimpse(re_pairs)
```

## Create Group Statistics Table

Extract one row per group with summary statistics for downstream analyses.
Combines paralog stats from `re_pairs` with allelic stats from `re_table`.

```{r group-stats}
# Extract unique allelic group statistics from re_table (one row per group)
allelic_stats <- re_table %>%
    distinct(
        cluster_id, sample, haplotype,
        n_paralogs, n_res_allelic,
        group_mean_allelic_diff, group_sd_allelic_diff, group_cv_allelic_diff
    ) %>%
    # Rename for clarity in group_stats
    rename(
        mean_allelic_diff = group_mean_allelic_diff,
        sd_allelic_diff = group_sd_allelic_diff,
        cv_allelic_diff = group_cv_allelic_diff
    )

# Extract unique paralog group statistics from re_pairs (one row per group)
paralog_stats <- re_pairs %>%
    distinct(
        cluster_id, sample, haplotype,
        n_pairs,
        group_mean_fire_diff, group_sd_fire_diff, group_median_fire_diff,
        group_min_fire_diff, group_max_fire_diff, group_range_fire_diff,
        group_mean_identity, group_median_identity, group_min_identity, group_max_identity
    )

# Combine paralog and allelic statistics
group_stats <- paralog_stats %>%
    left_join(allelic_stats, by = c("cluster_id", "sample", "haplotype")) %>%
    # Add CV for paralog (SD / mean pairwise diff)
    mutate(
        cv_fire_diff = group_sd_fire_diff / group_mean_fire_diff
    ) %>%
    filter(!is.na(group_sd_fire_diff), !is.na(group_mean_fire_diff))

cat("\n=== Group Statistics ===\n")
cat("Total groups:", nrow(group_stats), "\n")
cat("Groups with identity data:", sum(!is.na(group_stats$group_median_identity)), "\n")
cat("Groups with allelic data:", sum(!is.na(group_stats$mean_allelic_diff)), "\n")

glimpse(group_stats)
```

## Save Data

Save the primary dataframes to RData for use in subsequent analyses.

```{r save-data}
save(
    re_table,
    re_pairs,
    group_stats,
    min_coverage,
    COLOR_PARALOG,
    COLOR_ALLELIC,
    FIGURE_DIR,
    file = glue("{DATA_DIR}/paralog_data.RData")
)

cat("\n=== Data saved to", glue("{DATA_DIR}/paralog_data.RData"), "===\n")
cat("Objects saved:\n")
cat("  - re_table:", nrow(re_table), "rows\n")
cat("  - re_pairs:", nrow(re_pairs), "rows\n")
cat("  - group_stats:", nrow(group_stats), "rows\n")
cat("  - min_coverage:", min_coverage, "\n")
cat("  - COLOR_PARALOG, COLOR_ALLELIC, FIGURE_DIR\n")
```

## Summary

```{r summary}
cat("=== Data Loading Summary ===\n\n")
cat("REs loaded:", nrow(re_table), "(after coverage + group filter)\n")
cat("Paralog pairs (all pairwise):", nrow(re_pairs), "\n")
cat("  - with alignment:", sum(!is.na(re_pairs$gap_compressed_identity)), "\n")
cat("  - without alignment:", sum(is.na(re_pairs$gap_compressed_identity)), "\n")
cat("Unique clusters:", n_distinct(re_table$cluster_id), "\n")
cat("Unique groups (cluster/sample/haplotype):", n_distinct(re_pairs$cluster_id, re_pairs$sample, re_pairs$haplotype), "\n")

cat("\n=== Pre-computed Statistics in re_table ===\n")
cat("Per-RE metrics:\n")
cat("  - allelic_diff: |fire_ratio_ref - fire_ratio_alt| at this RE\n")
cat("  - has_allelic_data: TRUE if sufficient alt coverage\n")
cat("Group-level metrics (same value for all REs in group):\n")
cat("  - n_paralogs: count of REs in group\n")
cat("  - n_res_allelic: count of REs with sufficient alt coverage\n")
cat("  - group_mean/sd/cv_allelic_diff: allelic difference stats\n")

cat("\n=== Pre-computed Statistics in re_pairs ===\n")
cat("Per-pair metrics:\n")
cat("  - fire_diff: absolute difference in FIRE ratio between paralogs\n")
cat("  - allelic_diff_query/target: ref vs alt FIRE difference for each RE\n")
cat("  - identity_bin: binned gap_compressed_identity\n")
cat("Group-level metrics (same value for all pairs in group):\n")
cat("  - n_pairs, n_paralogs: count of pairs and unique REs in group\n")
cat("  - group_mean/sd/median/min/max_fire_diff: FIRE difference stats\n")
cat("  - group_mean/median/min/max_identity: sequence identity stats\n")

cat("\n=== Pre-computed Statistics in group_stats ===\n")
cat("Paralog metrics (from re_pairs):\n")
cat("  - n_pairs, n_paralogs, cv_fire_diff\n")
cat("  - group_mean/sd/median/min/max/range_fire_diff\n")
cat("  - group_mean/median/min/max_identity\n")
cat("Allelic metrics (from re_table):\n")
cat("  - n_res_allelic: count of REs with sufficient alt coverage\n")
cat("  - mean_allelic_diff, sd_allelic_diff, cv_allelic_diff\n")
```

## Session Info

```{r session-info}
sessionInfo()
```
