---
title: "Data Loader - Paralog Chromatin Accessibility Analysis"
author: "PARE Pipeline"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Overview

This script loads and prepares the primary dataframes for the paralog chromatin accessibility analysis.

**Output dataframes saved to RData:**
- `re_table_filtered` - REs with cluster assignments and FIRE ratios (coverage filtered, groups with >=2 REs)
- `re_pairs` - Paralog pairs with FIRE data for both REs

## Load Libraries

```{r libraries}
library(tidyverse)
library(ggplot2)
library(glue)
library(data.table)
library(hexbin)
library(scales)

# Install mrvplot if not available
if (!requireNamespace("mrvplot", quietly = TRUE)) {
    message("Installing mrvplot from GitHub...")
    devtools::install_github("mrvollger/mrvplot")
}
library(mrvplot)

# Use mrvplot theme
theme_set(mrv_grid())

# Color palette for paralog vs allelic comparisons
COLOR_PARALOG <- "darkred"
COLOR_ALLELIC <- "gray40"

# Create output directory for figures
FIGURE_DIR <- "figures"
if (!dir.exists(FIGURE_DIR)) {
    dir.create(FIGURE_DIR, recursive = TRUE)
}

# Create output directory for data
DATA_DIR <- "data"
if (!dir.exists(DATA_DIR)) {
    dir.create(DATA_DIR, recursive = TRUE)
}

# Minimum coverage threshold
min_coverage <- 10
```

## Load RE Table

Load the RE table with cluster assignments and peak/FIRE information,
then calculate FIRE ratios and filter for sufficient coverage.

```{r load-re-table}
# Load REs with cluster assignments and peak information
re_table <- read_tsv(
    "results/pileup/peaks_with_clusters.bed.gz",
    col_types = cols(.default = col_guess())
) %>%
    rename(chrom = `#chrom`) %>%
    mutate(
        # Calculate FIRE ratio (proportion of fibers with accessible chromatin)
        fire_ratio_ref = fire_coverage_ref / coverage_ref,
        fire_ratio_alt = fire_coverage_alt / coverage_alt,
        # Handle division by zero
        fire_ratio_ref = if_else(coverage_ref == 0, NA_real_, fire_ratio_ref),
        fire_ratio_alt = if_else(coverage_alt == 0, NA_real_, fire_ratio_alt)
    )

cat("Loaded", nrow(re_table), "REs\n")
cat("Unique clusters:", n_distinct(re_table$cluster_id), "\n")
cat("Samples:", paste(unique(re_table$sample), collapse = ", "), "\n")

# Filter for sufficient coverage
re_table_coverage_filtered <- re_table %>%
    filter(coverage_ref >= min_coverage)

cat("REs with coverage >=", min_coverage, ":", nrow(re_table_coverage_filtered), "\n")
cat("REs filtered out:", nrow(re_table) - nrow(re_table_coverage_filtered), "\n")

# Filter to keep only groups (cluster, sample, haplotype) with at least 2 REs
# This is essential for paralog analysis - we need at least 2 REs to compare
groups_with_multiple_res <- re_table_coverage_filtered %>%
    count(cluster_id, sample, haplotype) %>%
    filter(n >= 2) %>%
    select(-n)

re_table_filtered <- re_table_coverage_filtered %>%
    semi_join(groups_with_multiple_res, by = c("cluster_id", "sample", "haplotype"))

n_singleton_groups <- nrow(re_table_coverage_filtered) - nrow(re_table_filtered)
cat("REs in singleton groups removed:", n_singleton_groups, "\n")
cat("REs after group filter:", nrow(re_table_filtered), "\n")

glimpse(re_table_filtered)
```

## Load RE Pairwise Table (Paralogs)

Load pairwise RE alignments from all-by-all analysis and filter for **paralog pairs** -
pairs where both REs are in the same cluster, same sample, and same haplotype.
Then join with RE table to add FIRE data and filter for pairs where both REs pass coverage thresholds.

```{r load-re-pairs}
# Load all pairwise RE alignments
all_pairs <- read_tsv(
    "results/all_by_all/re_pairwise_identity.tsv.gz",
    col_types = cols(
        query_re_id = col_character(),
        target_re_id = col_character(),
        query_cluster = col_character(),
        target_cluster = col_character(),
        same_cluster = col_logical(),
        query_sample = col_character(),
        target_sample = col_character(),
        query_haplotype = col_character(),
        target_haplotype = col_character(),
        .default = col_guess()
    )
)

cat("Total pairwise alignments:", nrow(all_pairs), "\n")

# Filter for paralog pairs: same cluster, same sample, same haplotype
re_pairs_raw <- all_pairs %>%
    filter(
        same_cluster == TRUE,
        query_sample == target_sample,
        query_haplotype == target_haplotype
    ) %>%
    rename(
        cluster_id = query_cluster,
        sample = query_sample,
        haplotype = query_haplotype
    ) %>%
    select(-target_cluster, -target_sample, -target_haplotype, -same_cluster)

cat("Paralog pairs (same cluster/sample/haplotype):", nrow(re_pairs_raw), "\n")

# Join RE metadata for both query and target REs
re_cols_to_join <- c("re_id", "fire_ratio_ref", "fire_ratio_alt", "coverage_ref", "coverage_alt")

re_pairs <- re_pairs_raw %>%
    # Join query RE info
    inner_join(
        re_table_filtered %>% select(all_of(re_cols_to_join)),
        by = c("query_re_id" = "re_id"),
        suffix = c("", "_query")
    ) %>%
    rename(
        fire_ratio_ref_query = fire_ratio_ref,
        fire_ratio_alt_query = fire_ratio_alt,
        coverage_ref_query = coverage_ref,
        coverage_alt_query = coverage_alt
    ) %>%
    # Join target RE info
    inner_join(
        re_table_filtered %>% select(all_of(re_cols_to_join)),
        by = c("target_re_id" = "re_id"),
        suffix = c("", "_target")
    ) %>%
    rename(
        fire_ratio_ref_target = fire_ratio_ref,
        fire_ratio_alt_target = fire_ratio_alt,
        coverage_ref_target = coverage_ref,
        coverage_alt_target = coverage_alt
    )

# Report filtering stats
n_before <- nrow(re_pairs_raw)
n_after <- nrow(re_pairs)
n_removed <- n_before - n_after
pct_removed <- round(100 * n_removed / n_before, 1)

cat("\n=== Filtering pairs for coverage ===\n")
cat("Pairs before filtering:", n_before, "\n")
cat("Pairs after filtering:", n_after, "\n")
cat("Pairs removed:", n_removed, "(", pct_removed, "%)\n")
cat("Clusters with paralogs:", n_distinct(re_pairs$cluster_id), "\n")

glimpse(re_pairs)
```

## Save Data

Save the primary dataframes to RData for use in subsequent analyses.

```{r save-data}
save(
    re_table_filtered,
    re_pairs,
    min_coverage,
    COLOR_PARALOG,
    COLOR_ALLELIC,
    FIGURE_DIR,
    file = glue("{DATA_DIR}/paralog_data.RData")
)

cat("\n=== Data saved to", glue("{DATA_DIR}/paralog_data.RData"), "===\n")
cat("Objects saved:\n")
cat("  - re_table_filtered:", nrow(re_table_filtered), "rows\n")
cat("  - re_pairs:", nrow(re_pairs), "rows\n")
cat("  - min_coverage:", min_coverage, "\n")
cat("  - COLOR_PARALOG, COLOR_ALLELIC, FIGURE_DIR\n")
```

## Summary

```{r summary}
cat("=== Data Loading Summary ===\n\n")
cat("REs loaded:", nrow(re_table_filtered), "(after coverage + group filter)\n")
cat("Paralog pairs:", nrow(re_pairs), "\n")
cat("Unique clusters:", n_distinct(re_table_filtered$cluster_id), "\n")
```

## Session Info

```{r session-info}
sessionInfo()
```
