---
title: "Paralog Within-Group Variance Analysis"
author: "PARE Pipeline"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Overview

This analysis explores the variance in chromatin accessibility (measured by FIRE score)
between paralogous regulatory elements within the same cluster, sample, and haplotype.

**Key question**: Do paralogs within the same genome show similar or divergent
chromatin accessibility patterns?

## Load Data

```{r load-data}
library(tidyverse)
library(ggplot2)
library(glue)
library(scales)

# Install mrvplot if not available
if (!requireNamespace("mrvplot", quietly = TRUE)) {
    devtools::install_github("mrvollger/mrvplot")
}
library(mrvplot)
theme_set(mrv_grid())

# Load pre-processed data from data loader
load("data/paralog_data.RData")

cat("Loaded data:\n")
cat("  - re_table:", nrow(re_table), "REs\n")
cat("  - re_pairs:", nrow(re_pairs), "paralog pairs\n")
cat("  - group_stats:", nrow(group_stats), "groups\n")
```

## Group Statistics Overview

The data loader provides `group_stats` with one row per group and pre-computed statistics.

```{r group-stats-overview}
glimpse(group_stats)

# Distribution of paralog counts
group_stats %>%
    count(n_paralogs) %>%
    ggplot(aes(x = n_paralogs, y = n)) +
    geom_col() +
    scale_x_continuous(breaks = seq(1, max(group_stats$n_paralogs), by = 1)) +
    labs(
        x = "Paralogs per group",
        y = "Count",
        title = "Paralog group sizes"
    )
mrv_ggsave(glue("{FIGURE_DIR}/paralog_group_sizes.pdf"), width = 3, height = 2)

cat("Paralog groups:", nrow(group_stats), "\n")
cat("Total paralogs in these groups:", sum(group_stats$n_paralogs), "\n")
```

## Variance Distribution

```{r variance-sd-distribution}
# Distribution of standard deviation within paralog groups
group_stats %>%
    ggplot(aes(x = group_sd_fire_diff)) +
    geom_histogram(bins = 50, fill = COLOR_PARALOG, alpha = 0.7) +
    labs(
        x = "SD of FIRE ratio",
        y = "Count",
        title = "Paralog variance distribution"
    )
mrv_ggsave(glue("{FIGURE_DIR}/fire_ratio_sd_distribution.pdf"), width = 3, height = 2)
```

```{r variance-range-distribution}
# Distribution of range
group_stats %>%
    ggplot(aes(x = group_range_fire_diff)) +
    geom_histogram(bins = 50, fill = COLOR_PARALOG, alpha = 0.7) +
    labs(
        x = "Range of FIRE ratio",
        y = "Count",
        title = "Paralog range distribution"
    )
mrv_ggsave(glue("{FIGURE_DIR}/fire_ratio_range_distribution.pdf"), width = 3, height = 2)
```

## Variance by Group Size

Does variance increase with number of paralogs?

```{r variance-by-size}
group_stats %>%
    ggplot(aes(x = factor(n_paralogs), y = group_sd_fire_diff)) +
    geom_boxplot(fill = "steelblue", alpha = 0.5) +
    geom_jitter(alpha = 0.3, width = 0.2, size = 0.5) +
    labs(
        x = "Paralogs in group",
        y = "SD of FIRE ratio",
        title = "Variance by paralog count"
    )
mrv_ggsave(glue("{FIGURE_DIR}/variance_by_paralog_count.pdf"), width = 3, height = 2)
```

## Coefficient of Variation

The coefficient of variation (CV = SD/mean) normalizes variance by the mean,
useful for comparing groups with different baseline accessibility.

```{r cv-distribution}
group_stats %>%
    filter(is.finite(cv_fire_diff), cv_fire_diff < 5) %>% # Remove extreme values
    ggplot(aes(x = cv_fire_diff)) +
    geom_histogram(bins = 50, fill = "purple", alpha = 0.7) +
    labs(
        x = "CV (SD/mean)",
        y = "Count",
        title = "Coefficient of variation"
    )
mrv_ggsave(glue("{FIGURE_DIR}/cv_distribution.pdf"), width = 3, height = 2)
```

## High vs Low Variance Groups

Identify paralog groups with unusually high or low variance.

```{r high-low-variance-categorize}
# Categorize groups by variance
median_sd <- median(group_stats$group_sd_fire_diff, na.rm = TRUE)

group_stats <- group_stats %>%
    mutate(
        variance_category = case_when(
            group_sd_fire_diff < median_sd / 2 ~ "Low variance",
            group_sd_fire_diff > median_sd * 2 ~ "High variance",
            TRUE ~ "Moderate variance"
        )
    )
```

```{r high-low-variance-plot}
group_stats %>%
    count(variance_category) %>%
    ggplot(aes(x = variance_category, y = n, fill = variance_category)) +
    geom_col() +
    labs(
        x = "Category",
        y = "Groups",
        title = "Variance categories"
    ) +
    theme(legend.position = "none")
mrv_ggsave(glue("{FIGURE_DIR}/variance_categories.pdf"), width = 3, height = 2)
```

```{r high-low-variance-table}
# Summary statistics by category
group_stats %>%
    group_by(variance_category) %>%
    summarise(
        n_groups = n(),
        mean_n_paralogs = mean(n_paralogs),
        mean_fire_diff = mean(group_mean_fire_diff, na.rm = TRUE),
        mean_identity = mean(group_median_identity, na.rm = TRUE),
        .groups = "drop"
    ) %>%
    knitr::kable(digits = 3)
```

## Example High Variance Groups

```{r high-variance-table}
# Top high-variance groups
high_var_groups <- group_stats %>%
    filter(variance_category == "High variance") %>%
    arrange(desc(group_sd_fire_diff)) %>%
    head(10)

high_var_groups %>%
    select(cluster_id, sample, haplotype, n_paralogs, group_sd_fire_diff, group_range_fire_diff) %>%
    knitr::kable(digits = 3, caption = "Top 10 high variance paralog groups")

# Get individual RE data for these groups
high_var_res <- re_table %>%
    inner_join(
        high_var_groups %>% select(cluster_id, sample, haplotype),
        by = c("cluster_id", "sample", "haplotype")
    )
```

```{r high-variance-plot}
# Plot individual REs in high variance groups
high_var_res %>%
    ggplot(aes(x = cluster_id, y = fire_ratio_ref, color = sample)) +
    geom_jitter(width = 0.2, size = 1, alpha = 0.7) +
    coord_flip() +
    labs(
        x = "Cluster",
        y = "FIRE ratio",
        title = "High variance groups",
        color = "Sample"
    ) +
    theme(axis.text.y = element_text(size = 6))
mrv_ggsave(glue("{FIGURE_DIR}/high_variance_examples.pdf"), width = 3, height = 2)
```

## Example Low Variance Groups

```{r low-variance-examples}
# Top low-variance groups (most consistent)
low_var_groups <- group_stats %>%
    filter(variance_category == "Low variance", n_paralogs >= 3) %>%
    arrange(group_sd_fire_diff) %>%
    head(10)

low_var_groups %>%
    select(cluster_id, sample, haplotype, n_paralogs, group_sd_fire_diff, group_range_fire_diff) %>%
    knitr::kable(digits = 3, caption = "Top 10 low variance paralog groups (>=3 REs)")
```

## Random Cluster Examples

Violin plots showing FIRE ratio distribution for randomly selected clusters with sufficient REs.

```{r random-cluster-violins}
set.seed(42)

# Select random clusters with at least 4 REs for better visualization
random_clusters <- group_stats %>%
    filter(n_paralogs >= 4) %>%
    slice_sample(n = 10) %>%
    select(cluster_id, sample, haplotype)

# Get individual RE data for these clusters
random_cluster_res <- re_table %>%
    inner_join(random_clusters, by = c("cluster_id", "sample", "haplotype")) %>%
    mutate(
        cluster_label = glue("{cluster_id}\n({sample}, {haplotype})")
    )

# Show cluster info (use pre-computed stats from group_stats)
group_stats %>%
    inner_join(random_clusters, by = c("cluster_id", "sample", "haplotype")) %>%
    select(cluster_id, sample, haplotype, n_paralogs, group_mean_fire_diff, group_sd_fire_diff) %>%
    knitr::kable(digits = 3, caption = "Random cluster summary")


# Violin plot with points, colored by implicit vs annotated peaks
random_cluster_res %>%
    mutate(is_implicit = str_starts(re_id, "IMPLICIT_")) %>%
    ggplot(aes(x = cluster_label, y = fire_ratio_ref)) +
    geom_violin(alpha = 0.3, scale = "width", fill = "gray") +
    geom_jitter(aes(color = is_implicit), width = 0.1, size = 1, alpha = 0.8) +
    scale_color_manual(
        values = c("FALSE" = "steelblue", "TRUE" = "firebrick"),
        labels = c("FALSE" = "Annotated", "TRUE" = "Implicit")
    ) +
    labs(
        x = "Cluster",
        y = "FIRE ratio",
        title = "Random clusters",
        color = "Type"
    ) +
    theme(
        axis.text.x = element_text(size = 5, angle = 45, hjust = 1),
        legend.key.size = unit(0.3, "cm")
    )
mrv_ggsave(glue("{FIGURE_DIR}/random_cluster_violins.pdf"), width = 3, height = 2)
```

## Summary Statistics

```{r summary}
cat("=== Summary Statistics ===\n\n")
cat("Total paralog groups analyzed:", nrow(group_stats), "\n")
cat("Total paralogs in analysis:", sum(group_stats$n_paralogs), "\n\n")

cat("FIRE ratio variance (SD) across paralog groups:\n")
summary(group_stats$group_sd_fire_diff)

cat("\nFIRE ratio range across paralog groups:\n")
summary(group_stats$group_range_fire_diff)

cat(
    "\nProportion of groups with high variance (>2x median):",
    mean(group_stats$variance_category == "High variance"), "\n"
)
cat(
    "Proportion of groups with low variance (<0.5x median):",
    mean(group_stats$variance_category == "Low variance"), "\n"
)
```

## Session Info

```{r session-info}
sessionInfo()
```
