---
title: "Paralog vs Allelic Variance Comparison"
author: "PARE Pipeline"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Overview

Compare variance between paralogs (within same sample/haplotype) to allelic variance
(ref vs alt at each individual RE position).

**Key insight**:
- **Paralog variance**: SD of FIRE ratios among REs on the *same* sample/haplotype within a cluster
- **Allelic variance**: Difference between `fire_ratio_ref` and `fire_ratio_alt` at each RE position (comparing the two haplotypes at the same genomic location)

## Load Data

```{r load-data}
library(tidyverse)
library(ggplot2)
library(glue)
library(scales)
library(ggpubr)

# Install mrvplot if not available
if (!requireNamespace("mrvplot", quietly = TRUE)) {
    devtools::install_github("mrvollger/mrvplot")
}
library(mrvplot)
theme_set(mrv_grid())

# Load pre-processed data from data loader
load("data/paralog_data.RData")

cat("Loaded data:\n")
cat("  - re_table:", nrow(re_table), "REs\n")
cat("  - re_pairs:", nrow(re_pairs), "paralog pairs\n")
cat("  - group_stats:", nrow(group_stats), "groups\n")
```

## Prepare Variance Comparison

Use pre-computed paralog and allelic statistics from `group_stats`.

```{r variance-comparison-prep}
# Filter for groups with both paralog and allelic data (requires >=2 REs with sufficient alt coverage)
variance_comparison <- group_stats %>%
    filter(
        !is.na(sd_allelic_diff),
        n_res_allelic >= 2
    ) %>%
    mutate(
        variance_ratio = group_sd_fire_diff / sd_allelic_diff,
        variance_ratio_log = log2(variance_ratio)
    )

cat("Groups with both paralog and allelic data:", nrow(variance_comparison), "\n")
cat("Groups excluded (insufficient allelic data):", nrow(group_stats) - nrow(variance_comparison), "\n")
```

## Paralog vs Allelic Variance Scatter

```{r variance-comparison-scatter}
# Compare paralog SD to allelic SD
variance_comparison %>%
    ggplot(aes(x = group_sd_fire_diff, y = sd_allelic_diff)) +
    geom_point(alpha = 0.5, size = 0.5) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
    stat_regline_equation(size = 2) +
    labs(
        x = "Paralog SD",
        y = "Allelic SD",
        title = "Paralog vs Allelic"
    ) +
    coord_fixed()
mrv_ggsave(glue("{FIGURE_DIR}/paralog_vs_allelic_scatter.pdf"), width = 3, height = 3)
```

## Variance Distribution Comparison

```{r variance-comparison-histogram}
# Distribution comparison
variance_long <- variance_comparison %>%
    select(cluster_id, sample, haplotype, group_sd_fire_diff, sd_allelic_diff) %>%
    pivot_longer(
        cols = c(group_sd_fire_diff, sd_allelic_diff),
        names_to = "variance_type",
        values_to = "variance"
    ) %>%
    mutate(
        variance_type = case_when(
            variance_type == "group_sd_fire_diff" ~ "Paralog",
            variance_type == "sd_allelic_diff" ~ "Allelic"
        )
    )

variance_long %>%
    ggplot(aes(x = variance, fill = variance_type)) +
    geom_histogram(bins = 50, alpha = 0.6, position = "identity") +
    scale_fill_manual(values = c("Paralog" = COLOR_PARALOG, "Allelic" = COLOR_ALLELIC)) +
    labs(
        x = "Variance (s.d. within paralog clusters)",
        y = "Count",
        fill = "Type",
        title = "Paralog vs Allelic"
    ) +
    theme(legend.key.size = unit(0.3, "cm"))

mrv_ggsave(glue("{FIGURE_DIR}/paralog_vs_allelic_hist.pdf"), width = 3, height = 2)
```

## Statistical Comparison

```{r variance-comparison-stats}
# Summary stats with medians for all three metrics
cat("\n=== Variance Comparison Summary ===\n")

# Mean (of pairwise differences)
cat("\n--- Mean |diff| ---\n")
cat(
    "Paralog mean |diff|:", round(median(variance_comparison$group_mean_fire_diff, na.rm = TRUE), 4),
    "(median),", round(mean(variance_comparison$group_mean_fire_diff, na.rm = TRUE), 4), "(mean)\n"
)
cat(
    "Allelic mean |diff|:", round(median(variance_comparison$mean_allelic_diff, na.rm = TRUE), 4),
    "(median),", round(mean(variance_comparison$mean_allelic_diff, na.rm = TRUE), 4), "(mean)\n"
)

# SD
cat("\n--- SD ---\n")
cat(
    "Paralog SD:", round(median(variance_comparison$group_sd_fire_diff, na.rm = TRUE), 4),
    "(median),", round(mean(variance_comparison$group_sd_fire_diff, na.rm = TRUE), 4), "(mean)\n"
)
cat(
    "Allelic SD:", round(median(variance_comparison$sd_allelic_diff, na.rm = TRUE), 4),
    "(median),", round(mean(variance_comparison$sd_allelic_diff, na.rm = TRUE), 4), "(mean)\n"
)

# CV (filter for finite values)
cv_data_temp <- variance_comparison %>%
    filter(is.finite(cv_fire_diff), is.finite(cv_allelic_diff))
cat("\n--- CV ---\n")
cat(
    "Paralog CV:", round(median(cv_data_temp$cv_fire_diff, na.rm = TRUE), 4),
    "(median),", round(mean(cv_data_temp$cv_fire_diff, na.rm = TRUE), 4), "(mean)\n"
)
cat(
    "Allelic CV:", round(median(cv_data_temp$cv_allelic_diff, na.rm = TRUE), 4),
    "(median),", round(mean(cv_data_temp$cv_allelic_diff, na.rm = TRUE), 4), "(mean)\n"
)

# Paired comparison for SD
cat("\n\nPaired t-test (paralog SD vs allelic SD):\n")
t_test <- t.test(
    variance_comparison$group_sd_fire_diff,
    variance_comparison$sd_allelic_diff,
    paired = TRUE
)
print(t_test)

# What proportion have higher paralog variance?
prop_paralog_higher <- mean(variance_comparison$group_sd_fire_diff > variance_comparison$sd_allelic_diff, na.rm = TRUE)
cat("\nProportion where paralog variance > allelic variance:", round(prop_paralog_higher, 3), "\n")
```

## Coefficient of Variation Comparison

The coefficient of variation (CV = SD/mean) normalizes variance by the mean, useful for comparing
groups with different baseline values.

```{r cv-comparison-stats}
# Filter for finite CV values
cv_data <- variance_comparison %>%
    filter(is.finite(cv_fire_diff), is.finite(cv_allelic_diff))

cat("\n=== CV Comparison Summary ===\n")
cat("\nParalog CV (SD/mean of FIRE ratio):\n")
summary(cv_data$cv_fire_diff)
cat("\nAllelic CV (SD/mean of |ref - alt|):\n")
summary(cv_data$cv_allelic_diff)

# Paired comparison
cat("\n\nPaired t-test (paralog CV vs allelic CV):\n")
t_test_cv <- t.test(
    cv_data$cv_fire_diff,
    cv_data$cv_allelic_diff,
    paired = TRUE
)
print(t_test_cv)

# What proportion have higher paralog CV?
prop_cv_paralog_higher <- mean(cv_data$cv_fire_diff > cv_data$cv_allelic_diff, na.rm = TRUE)
cat("\nProportion where paralog CV > allelic CV:", round(prop_cv_paralog_higher, 3), "\n")
```

```{r cv-comparison-scatter}
# Compare paralog CV to allelic CV
cv_data %>%
    filter(cv_fire_diff < 5, cv_allelic_diff < 5) %>% # Remove extreme outliers
    ggplot(aes(x = cv_fire_diff, y = cv_allelic_diff)) +
    geom_point(alpha = 0.5, size = 0.5) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
    stat_regline_equation(size = 2) +
    labs(
        x = "Paralog CV",
        y = "Allelic CV",
        title = "Paralog vs Allelic CV"
    ) +
    coord_fixed()
mrv_ggsave(glue("{FIGURE_DIR}/paralog_vs_allelic_cv_scatter.pdf"), width = 3, height = 3)
```

```{r cv-comparison-histogram}
# Distribution comparison of CV
cv_long <- cv_data %>%
    filter(cv_fire_diff < 5, cv_allelic_diff < 5) %>%
    select(cluster_id, sample, haplotype, cv_fire_diff, cv_allelic_diff) %>%
    pivot_longer(
        cols = c(cv_fire_diff, cv_allelic_diff),
        names_to = "cv_type",
        values_to = "cv"
    ) %>%
    mutate(
        cv_type = case_when(
            cv_type == "cv_fire_diff" ~ "Paralog",
            cv_type == "cv_allelic_diff" ~ "Allelic"
        )
    )

cv_long %>%
    ggplot(aes(x = cv, fill = cv_type)) +
    geom_histogram(bins = 50, alpha = 0.6, position = "identity") +
    scale_fill_manual(values = c("Paralog" = COLOR_PARALOG, "Allelic" = COLOR_ALLELIC)) +
    labs(
        x = "Coefficient of Variation (SD/mean)",
        y = "Count",
        fill = "Type",
        title = "Paralog vs Allelic CV"
    ) +
    theme(legend.key.size = unit(0.3, "cm"))

mrv_ggsave(glue("{FIGURE_DIR}/paralog_vs_allelic_cv_hist.pdf"), width = 3, height = 2)
```

## Variance Ratio Distribution

```{r variance-ratio-plot}
# Distribution of variance ratio
variance_comparison %>%
    filter(is.finite(variance_ratio_log)) %>%
    ggplot(aes(x = variance_ratio_log)) +
    geom_histogram(bins = 50, fill = "steelblue", alpha = 0.7) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
    labs(
        x = "log2(Paralog s.d. / Allelic s.d.)",
        y = "Count",
        title = "Ratio of Paralog to Allelic Variance",
        subtitle = "Positive = more paralog variance, Negative = more allelic variance"
    )
mrv_ggsave(glue("{FIGURE_DIR}/variance_ratio_distribution.pdf"), width = 3, height = 2)
```

```{r variance-ratio-summary}
cat("\nlog2(paralog/allelic) ratio:\n")
summary(variance_comparison$variance_ratio_log[is.finite(variance_comparison$variance_ratio_log)])
```

## Stratified by Sequence Identity

To understand how sequence identity affects the comparison between paralog and allelic variance,
we stratify the analysis by the median pairwise sequence identity within each cluster.

```{r variance-by-identity-prep}
# Use pre-computed identity stats from group_stats (already joined via variance_comparison)
# Add identity bins based on median identity
variance_by_identity <- variance_comparison %>%
    mutate(
        identity_bin = cut(
            group_median_identity,
            breaks = c(0, 0.90, 0.95, 0.98, 0.99, 0.999, 1.0),
            labels = c("<90%", "90-95%", "95-98%", "98-99%", "99-99.9%", ">99.9%"),
            include.lowest = TRUE
        )
    )

cat("Clusters with variance and identity data:", nrow(variance_by_identity), "\n")
cat("\nClusters per identity bin:\n")
table(variance_by_identity$identity_bin)
```

```{r variance-by-identity-violin}
# Prepare data in long format for violin plot with all three stat types
variance_by_identity_long <- variance_by_identity %>%
    filter(!is.na(identity_bin)) %>%
    filter(is.finite(cv_fire_diff), is.finite(cv_allelic_diff)) %>%
    select(
        cluster_id, sample, haplotype, identity_bin,
        group_mean_fire_diff, mean_allelic_diff,
        group_sd_fire_diff, sd_allelic_diff,
        cv_fire_diff, cv_allelic_diff
    ) %>%
    pivot_longer(
        cols = c(
            group_mean_fire_diff, mean_allelic_diff,
            group_sd_fire_diff, sd_allelic_diff,
            cv_fire_diff, cv_allelic_diff
        ),
        names_to = "metric",
        values_to = "value"
    ) %>%
    mutate(
        stat_type = case_when(
            str_detect(metric, "mean") ~ "Mean |diff|",
            str_detect(metric, "sd") ~ "SD",
            str_detect(metric, "cv") ~ "CV"
        ),
        stat_type = factor(stat_type, levels = c("Mean |diff|", "SD", "CV")),
        group_type = case_when(
            str_detect(metric, "fire_diff") ~ "Paralog",
            str_detect(metric, "allelic_diff$") ~ "Allelic"
        ),
        group_type = factor(group_type, levels = c("Paralog", "Allelic"))
    )

# Calculate stats for annotations (per identity bin and stat type)
# Include median for each group type
variance_identity_bin_stats <- variance_by_identity_long %>%
    group_by(identity_bin, stat_type, group_type) %>%
    summarise(
        median_val = median(value, na.rm = TRUE),
        .groups = "drop"
    ) %>%
    pivot_wider(
        names_from = group_type,
        values_from = median_val,
        names_prefix = "med_"
    ) %>%
    left_join(
        variance_by_identity_long %>%
            group_by(identity_bin, stat_type) %>%
            summarise(
                n_clusters = n_distinct(paste(cluster_id, sample, haplotype)),
                y_pos = max(value, na.rm = TRUE) * 1.05,
                .groups = "drop"
            ),
        by = c("identity_bin", "stat_type")
    )

# Violin plot comparing paralog vs allelic by identity bin, faceted by stat type
variance_by_identity_long %>%
    ggplot(aes(x = identity_bin, y = value, fill = group_type)) +
    geom_violin(alpha = 0.7, scale = "width", position = position_dodge(0.8)) +
    geom_boxplot(width = 0.15, position = position_dodge(0.8), outlier.shape = NA, alpha = 0.8) +
    geom_text(
        data = variance_identity_bin_stats,
        aes(
            x = identity_bin, y = y_pos,
            label = glue("n={scales::comma(n_clusters)}\nP:{round(med_Paralog, 2)} A:{round(med_Allelic, 2)}")
        ),
        size = 1.8, vjust = 0, inherit.aes = FALSE
    ) +
    facet_wrap(~stat_type, scales = "free_y", ncol = 1) +
    scale_fill_manual(values = c("Paralog" = COLOR_PARALOG, "Allelic" = COLOR_ALLELIC)) +
    scale_y_continuous(expand = expansion(mult = c(0.05, 0.20))) +
    labs(
        x = "Median cluster sequence identity",
        y = "Value",
        fill = "Type",
        title = "Paralog vs Allelic by sequence identity",
        caption = "n = clusters, P = Paralog median, A = Allelic median"
    ) +
    theme(
        axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "top",
        legend.key.size = unit(0.4, "cm"),
        strip.text = element_text(face = "bold")
    )
mrv_ggsave(glue("{FIGURE_DIR}/variance_by_identity_violin.pdf"), width = 4, height = 4)
```

## Ratio by Identity (Mean, SD, CV)

```{r ratio-by-identity}
# Calculate ratios for all three stat types
variance_by_identity <- variance_by_identity %>%
    mutate(
        mean_ratio_log = log2(group_mean_fire_diff / mean_allelic_diff),
        sd_ratio_log = log2(group_sd_fire_diff / sd_allelic_diff),
        cv_ratio_log = log2(cv_fire_diff / cv_allelic_diff)
    )

# Prepare long format for faceted plot
ratio_by_identity_long <- variance_by_identity %>%
    filter(!is.na(identity_bin)) %>%
    filter(
        is.finite(mean_ratio_log),
        is.finite(sd_ratio_log),
        is.finite(cv_ratio_log)
    ) %>%
    select(
        cluster_id, sample, haplotype, identity_bin,
        mean_ratio_log, sd_ratio_log, cv_ratio_log
    ) %>%
    pivot_longer(
        cols = c(mean_ratio_log, sd_ratio_log, cv_ratio_log),
        names_to = "ratio_type",
        values_to = "ratio_log"
    ) %>%
    mutate(
        stat_type = case_when(
            ratio_type == "mean_ratio_log" ~ "Mean |diff|",
            ratio_type == "sd_ratio_log" ~ "SD",
            ratio_type == "cv_ratio_log" ~ "CV"
        ),
        stat_type = factor(stat_type, levels = c("Mean |diff|", "SD", "CV"))
    )

# Calculate stats for annotations
ratio_stats <- ratio_by_identity_long %>%
    group_by(identity_bin, stat_type) %>%
    summarise(
        n = n_distinct(paste(cluster_id, sample, haplotype)),
        pct_paralog_higher = mean(ratio_log > 0, na.rm = TRUE) * 100,
        y_pos = max(ratio_log, na.rm = TRUE) * 1.05,
        .groups = "drop"
    )

# Faceted violin plot
ratio_by_identity_long %>%
    ggplot(aes(x = identity_bin, y = ratio_log, fill = identity_bin)) +
    geom_violin(alpha = 0.7, scale = "width") +
    geom_boxplot(width = 0.1, fill = "white", outlier.shape = NA) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
    geom_text(
        data = ratio_stats,
        aes(
            x = identity_bin, y = y_pos,
            label = glue("n={scales::comma(n)}\n{round(pct_paralog_higher)}%")
        ),
        size = 2, vjust = 0, inherit.aes = FALSE
    ) +
    facet_wrap(~stat_type, scales = "free_y", ncol = 1) +
    scale_fill_viridis_d(option = "plasma", direction = -1) +
    scale_y_continuous(expand = expansion(mult = c(0.05, 0.18))) +
    labs(
        x = "Median cluster sequence identity",
        y = "log2(Paralog / Allelic)",
        title = "Ratio by sequence identity",
        caption = "% = clusters where paralog > allelic"
    ) +
    theme(
        legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1),
        strip.text = element_text(face = "bold")
    )
mrv_ggsave(glue("{FIGURE_DIR}/ratio_by_identity.pdf"), width = 4, height = 4)
```

## Statistical Tests by Identity

```{r ratio-by-identity-stats}
# Summary statistics table for all three stat types
ratio_summary_stats <- ratio_by_identity_long %>%
    group_by(stat_type, identity_bin) %>%
    summarise(
        n = n_distinct(paste(cluster_id, sample, haplotype)),
        median_ratio = median(ratio_log, na.rm = TRUE),
        mean_ratio = mean(ratio_log, na.rm = TRUE),
        pct_paralog_higher = mean(ratio_log > 0, na.rm = TRUE) * 100,
        .groups = "drop"
    )

ratio_summary_stats %>%
    mutate(pct_paralog_higher = round(pct_paralog_higher, 1)) %>%
    knitr::kable(
        digits = 3,
        caption = "Ratio statistics by sequence identity (all metrics)",
        col.names = c(
            "Metric", "Identity bin", "N clusters", "Median log2 ratio",
            "Mean log2 ratio", "% Paralog > Allelic"
        )
    )

# Statistical tests: does ratio differ by identity for each stat type?
cat("\n=== Kruskal-Wallis tests: ratio ~ identity bin ===\n")

for (stat in c("Mean |diff|", "SD", "CV")) {
    cat("\n---", stat, "ratio ---\n")
    test_data <- ratio_by_identity_long %>%
        filter(stat_type == stat, is.finite(ratio_log))
    kruskal_result <- kruskal.test(ratio_log ~ identity_bin, data = test_data)
    cat(
        "Chi-squared:", round(kruskal_result$statistic, 2),
        ", df:", kruskal_result$parameter,
        ", p-value:", format(kruskal_result$p.value, scientific = TRUE, digits = 3), "\n"
    )
}
```

## Session Info

```{r session-info}
sessionInfo()
```
