---
title: "Paralog Actuation Difference Stratified by Sequence Identity"
author: "PARE Pipeline"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Overview

Analyze how the difference in actuation (percent FIRE) between paralog pairs relates to their
sequence identity (gap-compressed identity from all-by-all alignments).

## Load Data

```{r load-data}
library(tidyverse)
library(ggplot2)
library(glue)
library(scales)
library(hexbin)
library(ggpubr)

# Install mrvplot if not available
if (!requireNamespace("mrvplot", quietly = TRUE)) {
    devtools::install_github("mrvollger/mrvplot")
}
library(mrvplot)
theme_set(mrv_grid())

# Load pre-processed data from data loader
load("data/paralog_data.RData")

cat("Loaded data:\n")
cat("  - re_table:", nrow(re_table), "REs\n")
cat("  - re_pairs:", nrow(re_pairs), "paralog pairs (all pairwise combinations)\n")
```

## Calculate Pairwise FIRE Difference

```{r pairwise-fire-diff}
# Use pre-computed fire_diff and identity_bin from data loader
# Filter for pairs with valid FIRE and identity data
re_pairs_with_fire_diff <- re_pairs %>%
    filter(
        !is.na(fire_diff),
        !is.na(gap_compressed_identity)
    )

cat("Paralog pairs with FIRE data:", nrow(re_pairs_with_fire_diff), "\n")
cat("\nPairs per identity bin:\n")
table(re_pairs_with_fire_diff$identity_bin)
```

## FIRE Difference by Identity - Violin Plot

```{r fire-diff-by-identity-violin}
# Calculate summary stats for annotations
bin_stats <- re_pairs_with_fire_diff %>%
    filter(!is.na(identity_bin)) %>%
    group_by(identity_bin) %>%
    summarise(
        n_pairs = n(),
        n_clusters = n_distinct(cluster_id),
        median_diff = median(fire_diff, na.rm = TRUE),
        y_pos = max(fire_diff, na.rm = TRUE) * 1.05,
        .groups = "drop"
    )

# Violin plot with n, cluster count, and median annotations
re_pairs_with_fire_diff %>%
    filter(!is.na(identity_bin)) %>%
    ggplot(aes(x = identity_bin, y = fire_diff, fill = identity_bin)) +
    geom_violin(alpha = 0.7, scale = "width") +
    geom_boxplot(width = 0.1, fill = "white", outlier.shape = NA) +
    geom_text(
        data = bin_stats,
        aes(
            x = identity_bin, y = y_pos,
            label = glue("n={scales::comma(n_pairs)}\nc={scales::comma(n_clusters)}\nmed={round(median_diff, 3)}")
        ),
        size = 2.5, vjust = 0, inherit.aes = FALSE
    ) +
    scale_fill_viridis_d(option = "plasma", direction = -1) +
    scale_y_continuous(expand = expansion(mult = c(0.05, 0.18))) +
    labs(
        x = "Gap-compressed sequence identity",
        y = "Percent actuation difference between paralogs",
        title = "Paralog actuation divergence by sequence identity",
        caption = "n = pairs, c = clusters, med = median |diff|"
    ) +
    theme(legend.position = "none")
mrv_ggsave(glue("{FIGURE_DIR}/fire_diff_by_identity_violin.pdf"), width = 4, height = 3)
```

## FIRE Difference vs Identity - Hexbin Plot

```{r fire-diff-by-identity-scatter}
# Hexbin plot of divergence (1 - identity) vs FIRE difference
# Using divergence on log10 scale spreads out high-identity pairs
# Cap at 0.0005 (99.95% identity) since we can't plot 100% on log scale
re_pairs_with_fire_diff %>%
    mutate(
        divergence = 1 - gap_compressed_identity,
        divergence = pmax(divergence, 0.0005) # Cap at 99.95% identity
    ) %>%
    ggplot(aes(x = divergence, y = fire_diff)) +
    geom_hex(bins = 50) +
    scale_fill_distiller("", palette = "Spectral", trans = "log10") +
    scale_x_continuous(
        trans = "log10",
        breaks = c(0.0005, 0.001, 0.01, 0.05, 0.1, 0.2),
        labels = c(">99.95%", "99.9%", "99%", "95%", "90%", "80%")
    ) +
    annotation_logticks(sides = "b") +
    geom_smooth(method = "lm", color = "darkred", se = FALSE, linewidth = 1) +
    stat_cor(method = "spearman", label.x.npc = "left", label.y.npc = "top", size = 3) +
    stat_regline_equation(label.x.npc = "left", label.y.npc = 0.85, size = 3) +
    labs(
        x = "Gap-compressed sequence identity (log10 divergence scale)",
        y = "Percent actuation difference between paralogs",
        title = "Paralog actuation divergence vs sequence identity"
    ) +
    mrv_grid() +
    theme(legend.position = "right")
mrv_ggsave(glue("{FIGURE_DIR}/fire_diff_vs_identity_hexbin.pdf"), width = 3, height = 2)
```

## Summary Statistics by Identity Bin

```{r fire-diff-by-identity-stats}
# Summary statistics by identity bin
fire_stats_by_identity <- re_pairs_with_fire_diff %>%
    filter(!is.na(identity_bin)) %>%
    group_by(identity_bin) %>%
    summarise(
        n_pairs = n(),
        mean_fire_diff = mean(fire_diff, na.rm = TRUE),
        median_fire_diff = median(fire_diff, na.rm = TRUE),
        sd_fire_diff = sd(fire_diff, na.rm = TRUE),
        pct_diff_gt_0.1 = mean(fire_diff > 0.1, na.rm = TRUE) * 100,
        pct_diff_gt_0.2 = mean(fire_diff > 0.2, na.rm = TRUE) * 100,
        .groups = "drop"
    )

fire_stats_by_identity %>%
    knitr::kable(
        digits = 3,
        caption = "Actuation difference statistics by sequence identity bin",
        col.names = c(
            "Identity", "N pairs", "Mean |diff|", "Median |diff|",
            "SD |diff|", "% diff > 10%", "% diff > 20%"
        )
    )
```

## Correlation Analysis

```{r fire-diff-correlation}
# Correlation between identity and FIRE difference
cor_test <- cor.test(
    re_pairs_with_fire_diff$gap_compressed_identity,
    re_pairs_with_fire_diff$fire_diff,
    method = "spearman"
)

cat("\n=== Correlation: Sequence Identity vs Actuation Difference ===\n")
cat("Spearman correlation:", round(cor_test$estimate, 4), "\n")
cat("p-value:", format(cor_test$p.value, scientific = TRUE, digits = 3), "\n")
```

## High Identity Pairs Analysis

```{r high-identity-low-fire-diff}
# What proportion of high-identity pairs (>98%) have similar FIRE ratios?
high_identity_pairs <- re_pairs_with_fire_diff %>%
    filter(gap_compressed_identity >= 0.98)

cat("\n=== High Identity Pairs (>=98% identity) ===\n")
cat("Total pairs:", nrow(high_identity_pairs), "\n")
cat("Mean actuation diff:", round(mean(high_identity_pairs$fire_diff, na.rm = TRUE), 3), "\n")
cat("Median actuation diff:", round(median(high_identity_pairs$fire_diff, na.rm = TRUE), 3), "\n")
cat("% with diff < 5%:", round(mean(high_identity_pairs$fire_diff < 0.05, na.rm = TRUE) * 100, 1), "%\n")
cat("% with diff < 10%:", round(mean(high_identity_pairs$fire_diff < 0.10, na.rm = TRUE) * 100, 1), "%\n")
cat("% with diff > 20%:", round(mean(high_identity_pairs$fire_diff > 0.20, na.rm = TRUE) * 100, 1), "%\n")
```

## Density Plot by Identity Bin

```{r fire-diff-density-by-identity}
# Density plot faceted by identity bin
re_pairs_with_fire_diff %>%
    filter(!is.na(identity_bin)) %>%
    ggplot(aes(x = fire_diff, fill = identity_bin)) +
    geom_density(alpha = 0.7) +
    facet_wrap(~identity_bin, ncol = 1, scales = "free_y") +
    scale_fill_viridis_d(option = "plasma", direction = -1) +
    labs(
        x = "Percent actuation difference between paralogs",
        y = "Density",
        title = "Distribution of actuation differences by identity"
    ) +
    theme(legend.position = "none")
mrv_ggsave(glue("{FIGURE_DIR}/fire_diff_density_by_identity.pdf"), width = 3, height = 3)
```

## Random Cluster Examples

Show individual paralog and allelic differences for randomly selected clusters.

```{r random-cluster-examples}
set.seed(42)

# Use re_pairs from data loader (all pairwise combinations)
# Filter for valid paralog differences
paralog_pairs_valid <- re_pairs %>%
    filter(!is.na(fire_diff))

# Select random groups with sufficient paralogs (>=3 REs means >=3 pairs)
random_groups <- paralog_pairs_valid %>%
    group_by(cluster_id, sample, haplotype) %>%
    summarise(n_pairs = n(), .groups = "drop") %>%
    filter(n_pairs >= 3) %>%
    slice_sample(n = 5)

# Get paralog differences for these groups
paralog_diffs <- paralog_pairs_valid %>%
    inner_join(random_groups %>% select(cluster_id, sample, haplotype),
        by = c("cluster_id", "sample", "haplotype")
    ) %>%
    transmute(
        cluster_id, sample, haplotype,
        pair_id = paste(query_re_id, target_re_id, sep = "_vs_"),
        fire_diff = fire_diff,
        diff_type = "Paralog"
    )

# Get allelic differences for unique REs in these groups
# Use pre-computed allelic_diff from re_table (already has abs())
allelic_diffs <- re_table %>%
    inner_join(random_groups %>% select(cluster_id, sample, haplotype),
        by = c("cluster_id", "sample", "haplotype")
    ) %>%
    filter(has_allelic_data) %>%
    transmute(
        cluster_id, sample, haplotype,
        pair_id = re_id,
        fire_diff = allelic_diff,
        diff_type = "Allelic"
    )

# Create labels and combine
paralog_diffs <- paralog_diffs %>%
    mutate(group_label = glue("{cluster_id}\n({sample}, {haplotype})"))
allelic_diffs <- allelic_diffs %>%
    mutate(group_label = glue("{cluster_id}\n({sample}, {haplotype})"))

combined_diffs <- bind_rows(paralog_diffs, allelic_diffs) %>%
    mutate(diff_type = factor(diff_type, levels = c("Paralog", "Allelic")))

# Summary table - show counts to verify pairwise vs per-RE
combined_diffs %>%
    group_by(group_label, diff_type) %>%
    summarise(
        n = n(),
        mean_diff = mean(fire_diff, na.rm = TRUE),
        median_diff = median(fire_diff, na.rm = TRUE),
        .groups = "drop"
    ) %>%
    pivot_wider(
        names_from = diff_type,
        values_from = c(n, mean_diff, median_diff)
    ) %>%
    knitr::kable(digits = 3, caption = "Random group summary (Paralog n = pairwise pairs, Allelic n = REs)")
```

```{r random-cluster-plot}
# Plot individual differences faceted by group
combined_diffs %>%
    ggplot(aes(x = diff_type, y = fire_diff, fill = diff_type)) +
    geom_violin(alpha = 0.5, scale = "width") +
    geom_jitter(width = 0.15, size = 1, alpha = 0.7) +
    facet_wrap(~group_label, nrow = 1) +
    scale_fill_manual(values = c("Paralog" = COLOR_PARALOG, "Allelic" = COLOR_ALLELIC)) +
    labs(
        x = "",
        y = "Actuation difference",
        title = "Paralog vs Allelic differences in random groups"
    ) +
    theme(
        legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1),
        strip.text = element_text(size = 5)
    )
mrv_ggsave(glue("{FIGURE_DIR}/random_cluster_paralog_allelic.pdf"), width = 5, height = 3)
```

## Session Info

```{r session-info}
sessionInfo()
```
