---
title: "Paralog FIRE Difference Stratified by Sequence Identity"
author: "PARE Pipeline"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Overview

Analyze how the difference in FIRE ratio between paralog pairs relates to their
sequence identity (gap-compressed identity from all-by-all alignments).

## Load Data

```{r load-data}
library(tidyverse)
library(ggplot2)
library(glue)
library(scales)
library(hexbin)
library(ggpubr)

# Install mrvplot if not available
if (!requireNamespace("mrvplot", quietly = TRUE)) {
    devtools::install_github("mrvollger/mrvplot")
}
library(mrvplot)
theme_set(mrv_grid())

# Load pre-processed data from data loader
load("data/paralog_data.RData")

cat("Loaded data:\n")
cat("  - re_pairs:", nrow(re_pairs), "paralog pairs\n")
```

## Calculate Pairwise FIRE Difference

```{r pairwise-fire-diff}
# Calculate pairwise FIRE difference for each paralog pair
# Using the re_pairs table which has FIRE ratios for both query and target REs
re_pairs_with_fire_diff <- re_pairs %>%
    filter(
        !is.na(fire_ratio_ref_query),
        !is.na(fire_ratio_ref_target)
    ) %>%
    mutate(
        fire_diff = abs(fire_ratio_ref_query - fire_ratio_ref_target),
        # Create identity bins
        identity_bin = cut(
            gap_compressed_identity,
            breaks = c(0, 0.90, 0.95, 0.98, 0.99, 0.999, 1.0),
            labels = c("<90%", "90-95%", "95-98%", "98-99%", "99-99.9%", ">99.9%"),
            include.lowest = TRUE
        )
    )

cat("Paralog pairs with FIRE data:", nrow(re_pairs_with_fire_diff), "\n")
cat("\nPairs per identity bin:\n")
table(re_pairs_with_fire_diff$identity_bin)
```

## FIRE Difference by Identity - Violin Plot

```{r fire-diff-by-identity-violin}
# Calculate summary stats for annotations
bin_stats <- re_pairs_with_fire_diff %>%
    filter(!is.na(identity_bin)) %>%
    group_by(identity_bin) %>%
    summarise(
        n_pairs = n(),
        n_clusters = n_distinct(cluster_id),
        median_diff = median(fire_diff, na.rm = TRUE),
        y_pos = max(fire_diff, na.rm = TRUE) * 1.05,
        .groups = "drop"
    )

# Violin plot with n, cluster count, and median annotations
re_pairs_with_fire_diff %>%
    filter(!is.na(identity_bin)) %>%
    ggplot(aes(x = identity_bin, y = fire_diff, fill = identity_bin)) +
    geom_violin(alpha = 0.7, scale = "width") +
    geom_boxplot(width = 0.1, fill = "white", outlier.shape = NA) +
    geom_text(
        data = bin_stats,
        aes(
            x = identity_bin, y = y_pos,
            label = glue("n={scales::comma(n_pairs)}\nc={scales::comma(n_clusters)}\nmed={round(median_diff, 3)}")
        ),
        size = 2.5, vjust = 0, inherit.aes = FALSE
    ) +
    scale_fill_viridis_d(option = "plasma", direction = -1) +
    scale_y_continuous(expand = expansion(mult = c(0.05, 0.18))) +
    labs(
        x = "Gap-compressed sequence identity",
        y = "|FIRE ratio difference|",
        title = "Paralog FIRE divergence by sequence identity",
        caption = "n = pairs, c = clusters, med = median |diff|"
    ) +
    theme(legend.position = "none")
mrv_ggsave(glue("{FIGURE_DIR}/fire_diff_by_identity_violin.pdf"), width = 4, height = 3.5)
```

## FIRE Difference vs Identity - Hexbin Plot

```{r fire-diff-by-identity-scatter}
# Hexbin plot of divergence (1 - identity) vs FIRE difference
# Using divergence on log10 scale spreads out high-identity pairs
# Cap at 0.0005 (99.95% identity) since we can't plot 100% on log scale
re_pairs_with_fire_diff %>%
    mutate(
        divergence = 1 - gap_compressed_identity,
        divergence = pmax(divergence, 0.0005) # Cap at 99.95% identity
    ) %>%
    ggplot(aes(x = divergence, y = fire_diff)) +
    geom_hex(bins = 50) +
    scale_fill_distiller("", palette = "Spectral", trans = "log10") +
    scale_x_continuous(
        trans = "log10",
        breaks = c(0.0005, 0.001, 0.01, 0.05, 0.1, 0.2),
        labels = c(">99.95%", "99.9%", "99%", "95%", "90%", "80%")
    ) +
    annotation_logticks(sides = "b") +
    geom_smooth(method = "lm", color = "darkred", se = FALSE, linewidth = 1) +
    stat_cor(method = "spearman", label.x.npc = "left", label.y.npc = "top", size = 3) +
    stat_regline_equation(label.x.npc = "left", label.y.npc = 0.85, size = 3) +
    labs(
        x = "Gap-compressed sequence identity (log10 divergence scale)",
        y = "|FIRE ratio difference|",
        title = "Paralog FIRE divergence vs sequence identity"
    ) +
    mrv_grid() +
    theme(legend.position = "right")
mrv_ggsave(glue("{FIGURE_DIR}/fire_diff_vs_identity_hexbin.pdf"), width = 5, height = 3)
```

## Summary Statistics by Identity Bin

```{r fire-diff-by-identity-stats}
# Summary statistics by identity bin
fire_stats_by_identity <- re_pairs_with_fire_diff %>%
    filter(!is.na(identity_bin)) %>%
    group_by(identity_bin) %>%
    summarise(
        n_pairs = n(),
        mean_fire_diff = mean(fire_diff, na.rm = TRUE),
        median_fire_diff = median(fire_diff, na.rm = TRUE),
        sd_fire_diff = sd(fire_diff, na.rm = TRUE),
        pct_diff_gt_0.1 = mean(fire_diff > 0.1, na.rm = TRUE) * 100,
        pct_diff_gt_0.2 = mean(fire_diff > 0.2, na.rm = TRUE) * 100,
        .groups = "drop"
    )

fire_stats_by_identity %>%
    knitr::kable(
        digits = 3,
        caption = "FIRE ratio difference statistics by sequence identity bin",
        col.names = c(
            "Identity", "N pairs", "Mean |diff|", "Median |diff|",
            "SD |diff|", "% diff > 0.1", "% diff > 0.2"
        )
    )
```

## Correlation Analysis

```{r fire-diff-correlation}
# Correlation between identity and FIRE difference
cor_test <- cor.test(
    re_pairs_with_fire_diff$gap_compressed_identity,
    re_pairs_with_fire_diff$fire_diff,
    method = "spearman"
)

cat("\n=== Correlation: Sequence Identity vs FIRE Difference ===\n")
cat("Spearman correlation:", round(cor_test$estimate, 4), "\n")
cat("p-value:", format(cor_test$p.value, scientific = TRUE, digits = 3), "\n")
```

## High Identity Pairs Analysis

```{r high-identity-low-fire-diff}
# What proportion of high-identity pairs (>98%) have similar FIRE ratios?
high_identity_pairs <- re_pairs_with_fire_diff %>%
    filter(gap_compressed_identity >= 0.98)

cat("\n=== High Identity Pairs (>=98% identity) ===\n")
cat("Total pairs:", nrow(high_identity_pairs), "\n")
cat("Mean FIRE diff:", round(mean(high_identity_pairs$fire_diff, na.rm = TRUE), 3), "\n")
cat("Median FIRE diff:", round(median(high_identity_pairs$fire_diff, na.rm = TRUE), 3), "\n")
cat("% with FIRE diff < 0.05:", round(mean(high_identity_pairs$fire_diff < 0.05, na.rm = TRUE) * 100, 1), "%\n")
cat("% with FIRE diff < 0.10:", round(mean(high_identity_pairs$fire_diff < 0.10, na.rm = TRUE) * 100, 1), "%\n")
cat("% with FIRE diff > 0.20:", round(mean(high_identity_pairs$fire_diff > 0.20, na.rm = TRUE) * 100, 1), "%\n")
```

## Density Plot by Identity Bin

```{r fire-diff-density-by-identity}
# Density plot faceted by identity bin
re_pairs_with_fire_diff %>%
    filter(!is.na(identity_bin)) %>%
    ggplot(aes(x = fire_diff, fill = identity_bin)) +
    geom_density(alpha = 0.7) +
    facet_wrap(~identity_bin, ncol = 1, scales = "free_y") +
    scale_fill_viridis_d(option = "plasma", direction = -1) +
    labs(
        x = "|FIRE ratio difference|",
        y = "Density",
        title = "Distribution of FIRE differences by identity"
    ) +
    theme(legend.position = "none")
mrv_ggsave(glue("{FIGURE_DIR}/fire_diff_density_by_identity.pdf"), width = 4, height = 5)
```

## Session Info

```{r session-info}
sessionInfo()
```
