---
title: "DEPRECATED - Paralog Chromatin Accessibility Analysis"
author: "PARE Pipeline"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## DEPRECATED

**This file is deprecated.** Please use the split analysis files instead:

- `00_data_loader.Rmd` - Load and prepare data
- `01_paralog_variance.Rmd` - Paralog within-group variance analysis
- `02_paralog_vs_allelic.Rmd` - Paralog vs allelic variance comparison
- `03_fire_diff_by_identity.Rmd` - FIRE difference stratified by sequence identity

## Overview

This analysis explores the variance in chromatin accessibility (measured by FIRE score)
between paralogous regulatory elements within the same cluster, sample, and haplotype.

**Key question**: Do paralogs within the same genome show similar or divergent
chromatin accessibility patterns?

## Load Libraries

```{r libraries}
library(tidyverse)
library(ggplot2)
library(glue)
library(data.table)
library(hexbin)
library(scales)

# Install mrvplot if not available
if (!requireNamespace("mrvplot", quietly = TRUE)) {
    message("Installing mrvplot from GitHub...")
    devtools::install_github("mrvollger/mrvplot")
}
library(mrvplot)

# Use mrvplot theme
theme_set(mrv_grid())

# Color palette for paralog vs allelic comparisons
COLOR_PARALOG <- "darkred"
COLOR_ALLELIC <- "gray40"

# Create output directory for figures
FIGURE_DIR <- "figures"
if (!dir.exists(FIGURE_DIR)) {
    dir.create(FIGURE_DIR, recursive = TRUE)
}
```

## Load Data

### RE Table

First, we load the RE table with cluster assignments and peak/FIRE information,
then calculate FIRE ratios and filter for sufficient coverage.

```{r load-re-table}
# Load REs with cluster assignments and peak information
re_table <- read_tsv(
    "results/pileup/peaks_with_clusters.bed.gz",
    col_types = cols(.default = col_guess())
) %>%
    rename(chrom = `#chrom`) %>%
    mutate(
        # Calculate FIRE ratio (proportion of fibers with accessible chromatin)
        fire_ratio_ref = fire_coverage_ref / coverage_ref,
        fire_ratio_alt = fire_coverage_alt / coverage_alt,
        # Handle division by zero
        fire_ratio_ref = if_else(coverage_ref == 0, NA_real_, fire_ratio_ref),
        fire_ratio_alt = if_else(coverage_alt == 0, NA_real_, fire_ratio_alt)
    )
mean(is.na(re_table$fire_ratio_ref)) # Check for NAs
mean(is.na(re_table$fire_ratio_alt)) # Check for NAs

cat("Loaded", nrow(re_table), "REs\n")
cat("Unique clusters:", n_distinct(re_table$cluster_id), "\n")
cat("Samples:", paste(unique(re_table$sample), collapse = ", "), "\n")

# Filter for sufficient coverage
min_coverage <- 10
re_table_filtered <- re_table %>%
    filter(coverage_ref >= min_coverage)

cat("REs with coverage >=", min_coverage, ":", nrow(re_table_filtered), "\n")
cat("REs filtered out:", nrow(re_table) - nrow(re_table_filtered), "\n")

glimpse(re_table_filtered)
```

### RE Pairwise Table (Paralogs)

Load pairwise RE alignments from all-by-all analysis and filter for **paralog pairs** -
pairs where both REs are in the same cluster, same sample, and same haplotype.
Then join with RE table to add FIRE data and filter for pairs where both REs pass coverage thresholds.

```{r load-re-pairs}
# Load all pairwise RE alignments
all_pairs <- read_tsv(
    "results/all_by_all/re_pairwise_identity.tsv.gz",
    col_types = cols(
        query_re_id = col_character(),
        target_re_id = col_character(),
        query_cluster = col_character(),
        target_cluster = col_character(),
        same_cluster = col_logical(),
        query_sample = col_character(),
        target_sample = col_character(),
        query_haplotype = col_character(),
        target_haplotype = col_character(),
        .default = col_guess()
    )
)

cat("Total pairwise alignments:", nrow(all_pairs), "\n")

# Filter for paralog pairs: same cluster, same sample, same haplotype
re_pairs_raw <- all_pairs %>%
    filter(
        same_cluster == TRUE,
        query_sample == target_sample,
        query_haplotype == target_haplotype
    ) %>%
    rename(
        cluster_id = query_cluster,
        sample = query_sample,
        haplotype = query_haplotype
    ) %>%
    select(-target_cluster, -target_sample, -target_haplotype, -same_cluster)

cat("Paralog pairs (same cluster/sample/haplotype):", nrow(re_pairs_raw), "\n")

# Get the set of REs that pass coverage filter
passing_res <- re_table_filtered$re_id

# Join RE metadata for both query and target REs
re_cols_to_join <- c("re_id", "fire_ratio_ref", "fire_ratio_alt", "coverage_ref", "coverage_alt")

re_pairs <- re_pairs_raw %>%
    # Join query RE info
    inner_join(
        re_table_filtered %>% select(all_of(re_cols_to_join)),
        by = c("query_re_id" = "re_id"),
        suffix = c("", "_query")
    ) %>%
    rename(
        fire_ratio_ref_query = fire_ratio_ref,
        fire_ratio_alt_query = fire_ratio_alt,
        coverage_ref_query = coverage_ref,
        coverage_alt_query = coverage_alt
    ) %>%
    # Join target RE info
    inner_join(
        re_table_filtered %>% select(all_of(re_cols_to_join)),
        by = c("target_re_id" = "re_id"),
        suffix = c("", "_target")
    ) %>%
    rename(
        fire_ratio_ref_target = fire_ratio_ref,
        fire_ratio_alt_target = fire_ratio_alt,
        coverage_ref_target = coverage_ref,
        coverage_alt_target = coverage_alt
    )

# Report filtering stats
n_before <- nrow(re_pairs_raw)
n_after <- nrow(re_pairs)
n_removed <- n_before - n_after
pct_removed <- round(100 * n_removed / n_before, 1)

cat("\n=== Filtering pairs for coverage ===\n")
cat("Pairs before filtering:", n_before, "\n")
cat("Pairs after filtering:", n_after, "\n")
cat("Pairs removed:", n_removed, "(", pct_removed, "%)\n")
cat("Clusters with paralogs:", n_distinct(re_pairs$cluster_id), "\n")

glimpse(re_pairs)
```

```{r re-pairs-summary}
# Summary of pairs per group
pairs_per_group <- re_pairs %>%
    count(sample, haplotype, cluster_id, name = "n_pairs")

cat("\nPairs per group summary:\n")
summary(pairs_per_group$n_pairs)

# Distribution of pair counts
pairs_per_group %>%
    count(n_pairs) %>%
    head(20) %>%
    ggplot(aes(x = n_pairs, y = n)) +
    geom_col(fill = "steelblue") +
    labs(
        x = "Number of paralog pairs per group",
        y = "Count of groups",
        title = "Distribution of paralog pairs per (sample, haplotype, cluster)"
    )
mrv_ggsave(glue("{FIGURE_DIR}/pairs_per_group_distribution.pdf"), width = 4, height = 3)
```

## Identify Paralog Groups

Filter to clusters that have multiple paralogs on the same sample and haplotype.

```{r paralog-groups}
# Count paralogs per cluster/sample/haplotype
paralog_counts <- re_table_filtered %>%
    group_by(cluster_id, sample, haplotype) %>%
    summarise(
        n_paralogs = n(),
        .groups = "drop"
    ) %>%
    filter(n_paralogs >= 2)

# Distribution of paralog counts
paralog_counts %>%
    count(n_paralogs) %>%
    ggplot(aes(x = n_paralogs, y = n)) +
    geom_col() +
    scale_x_continuous(breaks = seq(1, max(paralog_counts$n_paralogs), by = 1)) +
    labs(
        x = "Paralogs per group",
        y = "Count",
        title = "Paralog group sizes"
    )
mrv_ggsave(glue("{FIGURE_DIR}/paralog_group_sizes.pdf"), width = 3, height = 2)

# Filter to groups with 2+ paralogs
paralog_groups <- paralog_counts
cat("Paralog groups (>=2 paralogs):", nrow(paralog_groups), "\n")
cat("Total paralogs in these groups:", sum(paralog_groups$n_paralogs), "\n")
```

## Calculate Within-Group Variance

For each paralog group, calculate the variance and range of FIRE ratios.

```{r within-group-variance}
# Join back to get the actual RE data for paralog groups
paralogs <- re_table_filtered %>%
    inner_join(paralog_groups, by = c("cluster_id", "sample", "haplotype"))

# Helper function to calculate mean of all pairwise absolute differences
mean_pairwise_diff <- function(x) {
    x <- x[!is.na(x)]
    if (length(x) < 2) {
        return(NA_real_)
    }
    # Get all pairwise combinations and compute mean absolute difference
    diffs <- combn(x, 2, function(pair) abs(pair[1] - pair[2]))
    mean(diffs)
}

# Calculate variance statistics per group
group_stats <- paralogs %>%
    group_by(cluster_id, sample, haplotype) %>%
    summarise(
        n_paralogs = n(),
        mean_fire_ratio = mean(fire_ratio_ref, na.rm = TRUE),
        sd_fire_ratio = sd(fire_ratio_ref, na.rm = TRUE),
        var_fire_ratio = var(fire_ratio_ref, na.rm = TRUE),
        min_fire_ratio = min(fire_ratio_ref, na.rm = TRUE),
        max_fire_ratio = max(fire_ratio_ref, na.rm = TRUE),
        range_fire_ratio = max_fire_ratio - min_fire_ratio,
        cv_fire_ratio = sd_fire_ratio / mean_fire_ratio, # coefficient of variation
        mean_fire_ratio_diff = mean_pairwise_diff(fire_ratio_ref), # mean of pairwise |diff|
        mean_coverage = mean(coverage_ref, na.rm = TRUE),
        .groups = "drop"
    ) %>%
    filter(!is.na(sd_fire_ratio))

glimpse(group_stats)
```

## Variance Distribution

```{r variance-sd-distribution}
# Distribution of standard deviation within paralog groups
group_stats %>%
    ggplot(aes(x = sd_fire_ratio)) +
    geom_histogram(bins = 50, fill = COLOR_PARALOG, alpha = 0.7) +
    labs(
        x = "SD of FIRE ratio",
        y = "Count",
        title = "Paralog variance distribution"
    )
mrv_ggsave(glue("{FIGURE_DIR}/fire_ratio_sd_distribution.pdf"), width = 3, height = 2)
```

```{r variance-range-distribution}
# Distribution of range
group_stats %>%
    ggplot(aes(x = range_fire_ratio)) +
    geom_histogram(bins = 50, fill = COLOR_PARALOG, alpha = 0.7) +
    labs(
        x = "Range of FIRE ratio",
        y = "Count",
        title = "Paralog range distribution"
    )
mrv_ggsave(glue("{FIGURE_DIR}/fire_ratio_range_distribution.pdf"), width = 3, height = 2)
```

## Variance by Group Size

Does variance increase with number of paralogs?

```{r variance-by-size}
group_stats %>%
    ggplot(aes(x = factor(n_paralogs), y = sd_fire_ratio)) +
    geom_boxplot(fill = "steelblue", alpha = 0.5) +
    geom_jitter(alpha = 0.3, width = 0.2, size = 0.5) +
    labs(
        x = "Paralogs in group",
        y = "SD of FIRE ratio",
        title = "Variance by paralog count"
    )
mrv_ggsave(glue("{FIGURE_DIR}/variance_by_paralog_count.pdf"), width = 3, height = 2)
```

## Coefficient of Variation

The coefficient of variation (CV = SD/mean) normalizes variance by the mean,
useful for comparing groups with different baseline accessibility.

```{r cv-distribution}
group_stats %>%
    filter(is.finite(cv_fire_ratio), cv_fire_ratio < 5) %>% # Remove extreme values
    ggplot(aes(x = cv_fire_ratio)) +
    geom_histogram(bins = 50, fill = "purple", alpha = 0.7) +
    labs(
        x = "CV (SD/mean)",
        y = "Count",
        title = "Coefficient of variation"
    )
mrv_ggsave(glue("{FIGURE_DIR}/cv_distribution.pdf"), width = 3, height = 2)
```

## High vs Low Variance Groups

Identify paralog groups with unusually high or low variance.

```{r high-low-variance-categorize}
# Categorize groups by variance
median_sd <- median(group_stats$sd_fire_ratio, na.rm = TRUE)

group_stats <- group_stats %>%
    mutate(
        variance_category = case_when(
            sd_fire_ratio < median_sd / 2 ~ "Low variance",
            sd_fire_ratio > median_sd * 2 ~ "High variance",
            TRUE ~ "Moderate variance"
        )
    )
```

```{r high-low-variance-plot}
group_stats %>%
    count(variance_category) %>%
    ggplot(aes(x = variance_category, y = n, fill = variance_category)) +
    geom_col() +
    labs(
        x = "Category",
        y = "Groups",
        title = "Variance categories"
    ) +
    theme(legend.position = "none")
mrv_ggsave(glue("{FIGURE_DIR}/variance_categories.pdf"), width = 3, height = 2)
```

```{r high-low-variance-table}
# Summary statistics by category
group_stats %>%
    group_by(variance_category) %>%
    summarise(
        n_groups = n(),
        mean_n_paralogs = mean(n_paralogs),
        mean_fire_ratio = mean(mean_fire_ratio, na.rm = TRUE),
        mean_coverage = mean(mean_coverage, na.rm = TRUE),
        .groups = "drop"
    ) %>%
    knitr::kable(digits = 3)
```

## Example High Variance Groups

```{r high-variance-table}
# Top high-variance groups
high_var_groups <- group_stats %>%
    filter(variance_category == "High variance") %>%
    arrange(desc(sd_fire_ratio)) %>%
    head(10)

high_var_groups %>%
    select(cluster_id, sample, haplotype, n_paralogs, mean_fire_ratio, sd_fire_ratio, range_fire_ratio) %>%
    knitr::kable(digits = 3, caption = "Top 10 high variance paralog groups")

# Get individual RE data for these groups
high_var_paralogs <- paralogs %>%
    inner_join(
        high_var_groups %>% select(cluster_id, sample, haplotype),
        by = c("cluster_id", "sample", "haplotype")
    )
```

```{r high-variance-plot}
# Plot individual paralogs in high variance groups
high_var_paralogs %>%
    ggplot(aes(x = cluster_id, y = fire_ratio_ref, color = sample)) +
    geom_jitter(width = 0.2, size = 1, alpha = 0.7) +
    coord_flip() +
    labs(
        x = "Cluster",
        y = "FIRE ratio",
        title = "High variance groups",
        color = "Sample"
    ) +
    theme(axis.text.y = element_text(size = 6))
mrv_ggsave(glue("{FIGURE_DIR}/high_variance_examples.pdf"), width = 3, height = 2)
```

## Example Low Variance Groups

```{r low-variance-examples}
# Top low-variance groups (most consistent)
low_var_groups <- group_stats %>%
    filter(variance_category == "Low variance", n_paralogs >= 3) %>%
    arrange(sd_fire_ratio) %>%
    head(10)

low_var_groups %>%
    select(cluster_id, sample, haplotype, n_paralogs, mean_fire_ratio, sd_fire_ratio, range_fire_ratio) %>%
    knitr::kable(digits = 3, caption = "Top 10 low variance paralog groups (>=3 paralogs)")
```

## Random Cluster Examples
Violin plots showing FIRE ratio distribution for 5 randomly selected clusters with sufficient paralogs.

```{r random-cluster-violins}
set.seed(42)

# Select 5 random clusters with at least 4 paralogs for better visualization
random_clusters <- group_stats %>%
    filter(n_paralogs >= 4) %>%
    slice_sample(n = 10) %>%
    select(cluster_id, sample, haplotype)

# Get individual RE data for these clusters
random_cluster_paralogs <- paralogs %>%
    inner_join(random_clusters, by = c("cluster_id", "sample", "haplotype")) %>%
    mutate(
        cluster_label = glue("{cluster_id}\n({sample}, {haplotype})")
    )

# Show cluster info
random_cluster_paralogs %>%
    group_by(cluster_id, sample, haplotype) %>%
    summarise(
        n_paralogs = n(),
        mean_fire_ratio = mean(fire_ratio_ref, na.rm = TRUE),
        sd_fire_ratio = sd(fire_ratio_ref, na.rm = TRUE),
        .groups = "drop"
    ) %>%
    knitr::kable(digits = 3, caption = "Random cluster summary")


# Violin plot with points, colored by implicit vs annotated peaks
random_cluster_paralogs %>%
    mutate(is_implicit = str_starts(re_id, "IMPLICIT_")) %>%
    ggplot(aes(x = cluster_label, y = fire_ratio_ref)) +
    geom_violin(alpha = 0.3, scale = "width", fill = "gray") +
    geom_jitter(aes(color = is_implicit), width = 0.1, size = 1, alpha = 0.8) +
    scale_color_manual(
        values = c("FALSE" = "steelblue", "TRUE" = "firebrick"),
        labels = c("FALSE" = "Annotated", "TRUE" = "Implicit")
    ) +
    labs(
        x = "Cluster",
        y = "FIRE ratio",
        title = "Random clusters",
        color = "Type"
    ) +
    theme(
        axis.text.x = element_text(size = 5, angle = 45, hjust = 1),
        legend.key.size = unit(0.3, "cm")
    )
mrv_ggsave(glue("{FIGURE_DIR}/random_cluster_violins.pdf"), width = 3, height = 2)
```

## Paralog vs Allelic Variance
Compare variance between paralogs (within same sample/haplotype) to allelic variance
(ref vs alt at each individual RE position).

**Key insight**:
- **Paralog variance**: SD of FIRE ratios among REs on the *same* sample/haplotype within a cluster
- **Allelic variance**: Difference between `fire_ratio_ref` and `fire_ratio_alt` at each RE position (comparing the two haplotypes at the same genomic location)

```{r paralog-vs-allelic}
# Calculate allelic difference at each RE position using the ref/alt columns
# These columns compare the coverage on the reference haplotype vs the alternate haplotype
# at each RE's genomic position
re_table_with_allelic <- re_table_filtered %>%
    filter(
        !is.na(fire_ratio_ref),
        !is.na(fire_ratio_alt),
        coverage_alt >= min_coverage # Require sufficient alt coverage too
    ) %>%
    mutate(
        allelic_diff = abs(fire_ratio_ref - fire_ratio_alt)
    )

cat("REs with sufficient ref AND alt coverage:", nrow(re_table_with_allelic), "\n")

# For a fair comparison, calculate the SD of allelic differences within each group
# This makes it comparable to sd_fire_ratio (SD of FIRE ratios across paralogs)
allelic_by_group <- re_table_with_allelic %>%
    group_by(cluster_id, sample, haplotype) %>%
    summarise(
        n_res_allelic = n(),
        mean_allelic_diff = mean(allelic_diff, na.rm = TRUE),
        sd_allelic_diff = sd(allelic_diff, na.rm = TRUE), # SD of allelic diffs within group
        .groups = "drop"
    ) %>%
    filter(n_res_allelic >= 2) # Need at least 2 REs to calculate SD

# Join with paralog variance stats
variance_comparison <- group_stats %>%
    inner_join(allelic_by_group, by = c("cluster_id", "sample", "haplotype"))
glimpse(variance_comparison)
cat("Groups with both paralog and allelic data:", nrow(variance_comparison), "\n")
```

```{r variance-comparison-scatter}
library(ggpubr)
# Compare paralog SD to mean allelic difference
variance_comparison %>%
    ggplot(aes(x = sd_fire_ratio, y = sd_allelic_diff)) +
    geom_point(alpha = 0.5, size = 0.5) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
    stat_regline_equation(size = 2) +
    labs(
        x = "Paralog SD",
        y = "Allelic |ref-alt|",
        title = "Paralog vs Allelic"
    ) +
    coord_fixed()
mrv_ggsave(glue("{FIGURE_DIR}/paralog_vs_allelic_scatter.pdf"), width = 3, height = 3)
```

```{r variance-comparison-histogram}
# Distribution comparison
variance_long <- variance_comparison %>%
    select(cluster_id, sample, haplotype, sd_fire_ratio, sd_allelic_diff) %>%
    pivot_longer(
        cols = c(sd_fire_ratio, sd_allelic_diff),
        names_to = "variance_type",
        values_to = "variance"
    ) %>%
    mutate(
        variance_type = case_when(
            variance_type == "sd_fire_ratio" ~ "Paralog",
            variance_type == "sd_allelic_diff" ~ "Allelic"
        )
    )

variance_long %>%
    ggplot(aes(x = variance, fill = variance_type)) +
    geom_histogram(bins = 50, alpha = 0.6, position = "identity") +
    scale_fill_manual(values = c("Paralog" = COLOR_PARALOG, "Allelic" = COLOR_ALLELIC)) +
    labs(
        x = "Variance (s.d. within paralog clusters)",
        y = "Count",
        fill = "Type",
        title = "Paralog vs Allelic"
    ) +
    theme(legend.key.size = unit(0.3, "cm"))

mrv_ggsave(glue("{FIGURE_DIR}/paralog_vs_allelic_hist.pdf"), width = 3, height = 2)
```

```{r variance-comparison-stats}
# Summary stats
cat("\n=== Variance Comparison Summary ===\n")
cat("\nParalog variance (SD of FIRE ratio within group):\n")
summary(variance_comparison$sd_fire_ratio)
cat("\nAllelic variance (SD of |ref - alt| within group):\n")
summary(variance_comparison$sd_allelic_diff)

# Paired comparison
cat("\n\nPaired t-test (paralog SD vs allelic SD):\n")
t_test <- t.test(
    variance_comparison$sd_fire_ratio,
    variance_comparison$sd_allelic_diff,
    paired = TRUE
)
print(t_test)

# What proportion have higher paralog variance?
prop_paralog_higher <- mean(variance_comparison$sd_fire_ratio > variance_comparison$sd_allelic_diff, na.rm = TRUE)
cat("\nProportion where paralog variance > allelic variance:", round(prop_paralog_higher, 3), "\n")
```

```{r variance-ratio-calc}
# Calculate ratio of paralog to allelic variance
variance_comparison <- variance_comparison %>%
    mutate(
        variance_ratio = sd_fire_ratio / sd_allelic_diff,
        variance_ratio_log = log2(variance_ratio)
    )
```

```{r variance-ratio-plot}
# Distribution of variance ratio
variance_comparison %>%
    filter(is.finite(variance_ratio_log)) %>%
    ggplot(aes(x = variance_ratio_log)) +
    geom_histogram(bins = 50, fill = "steelblue", alpha = 0.7) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
    labs(
        x = "log2(Paralog s.d. / Allelic s.d.)",
        y = "Count",
        title = "Ratio of Paralog to Allelic Variance",
        subtitle = "Positive = more paralog variance, Negative = more allelic variance"
    )
mrv_ggsave(glue("{FIGURE_DIR}/variance_ratio_distribution.pdf"), width = 3, height = 2)
```

```{r variance-ratio-summary}
cat("\nlog2(paralog/allelic) ratio:\n")
summary(variance_comparison$variance_ratio_log[is.finite(variance_comparison$variance_ratio_log)])
```

### Paralog vs Allelic Variance Stratified by Sequence Identity

To understand how sequence identity affects the comparison between paralog and allelic variance,
we stratify the analysis by the median pairwise sequence identity within each cluster.

```{r variance-by-identity-prep}
# Calculate median identity per cluster from pairwise alignments
cluster_identity <- re_pairs %>%
    group_by(cluster_id, sample, haplotype) %>%
    summarise(
        median_identity = median(gap_compressed_identity, na.rm = TRUE),
        mean_identity = mean(gap_compressed_identity, na.rm = TRUE),
        min_identity = min(gap_compressed_identity, na.rm = TRUE),
        max_identity = max(gap_compressed_identity, na.rm = TRUE),
        n_pairs = n(),
        .groups = "drop"
    )

# Join with variance comparison data
variance_by_identity <- variance_comparison %>%
    inner_join(cluster_identity, by = c("cluster_id", "sample", "haplotype")) %>%
    mutate(
        identity_bin = cut(
            median_identity,
            breaks = c(0, 0.90, 0.95, 0.98, 0.99, 0.999, 1.0),
            labels = c("<90%", "90-95%", "95-98%", "98-99%", "99-99.9%", ">99.9%"),
            include.lowest = TRUE
        )
    )

cat("Clusters with variance and identity data:", nrow(variance_by_identity), "\n")
cat("\nClusters per identity bin:\n")
table(variance_by_identity$identity_bin)
```

```{r variance-by-identity-violin}
# Prepare data in long format for violin plot
variance_by_identity_long <- variance_by_identity %>%
    filter(!is.na(identity_bin)) %>%
    select(cluster_id, sample, haplotype, identity_bin, sd_fire_ratio, sd_allelic_diff) %>%
    pivot_longer(
        cols = c(sd_fire_ratio, sd_allelic_diff),
        names_to = "variance_type",
        values_to = "variance"
    ) %>%
    mutate(
        variance_type = case_when(
            variance_type == "sd_fire_ratio" ~ "Paralog",
            variance_type == "sd_allelic_diff" ~ "Allelic"
        ),
        variance_type = factor(variance_type, levels = c("Paralog", "Allelic"))
    )

# Calculate stats for annotations (per identity bin, aggregated across variance types)
variance_identity_bin_stats <- variance_by_identity %>%
    filter(!is.na(identity_bin)) %>%
    group_by(identity_bin) %>%
    summarise(
        n_clusters = n(),
        median_paralog = median(sd_fire_ratio, na.rm = TRUE),
        median_allelic = median(sd_allelic_diff, na.rm = TRUE),
        y_pos = max(c(sd_fire_ratio, sd_allelic_diff), na.rm = TRUE) * 1.05,
        .groups = "drop"
    )

# Violin plot comparing paralog vs allelic variance by identity bin
variance_by_identity_long %>%
    ggplot(aes(x = identity_bin, y = variance, fill = variance_type)) +
    geom_violin(alpha = 0.7, scale = "width", position = position_dodge(0.8)) +
    geom_boxplot(width = 0.15, position = position_dodge(0.8), outlier.shape = NA, alpha = 0.8) +
    geom_text(
        data = variance_identity_bin_stats,
        aes(
            x = identity_bin, y = y_pos,
            label = glue("n={scales::comma(n_clusters)}")
        ),
        size = 2.5, vjust = 0, inherit.aes = FALSE
    ) +
    scale_fill_manual(values = c("Paralog" = COLOR_PARALOG, "Allelic" = COLOR_ALLELIC)) +
    scale_y_continuous(expand = expansion(mult = c(0.05, 0.15))) +
    labs(
        x = "Median cluster sequence identity",
        y = "SD of FIRE ratio",
        fill = "Type",
        title = "Paralog vs Allelic variance by sequence identity",
        caption = "n = number of clusters"
    ) +
    theme(
        axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "top",
        legend.key.size = unit(0.4, "cm")
    )
mrv_ggsave(glue("{FIGURE_DIR}/variance_by_identity_violin.pdf"), width = 5, height = 3.5)
```

```{r variance-ratio-by-identity}
# Plot variance ratio (paralog/allelic) stratified by identity
variance_ratio_stats <- variance_by_identity %>%
    filter(!is.na(identity_bin), is.finite(variance_ratio_log)) %>%
    group_by(identity_bin) %>%
    summarise(
        n = n(),
        median_ratio = median(variance_ratio_log, na.rm = TRUE),
        mean_ratio = mean(variance_ratio_log, na.rm = TRUE),
        pct_paralog_higher = mean(variance_ratio_log > 0, na.rm = TRUE) * 100,
        y_pos = max(variance_ratio_log, na.rm = TRUE) * 1.05,
        .groups = "drop"
    )

variance_by_identity %>%
    filter(!is.na(identity_bin), is.finite(variance_ratio_log)) %>%
    ggplot(aes(x = identity_bin, y = variance_ratio_log, fill = identity_bin)) +
    geom_violin(alpha = 0.7, scale = "width") +
    geom_boxplot(width = 0.1, fill = "white", outlier.shape = NA) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
    geom_text(
        data = variance_ratio_stats,
        aes(
            x = identity_bin, y = y_pos,
            label = glue("n={scales::comma(n)}\n{round(pct_paralog_higher)}% P>A")
        ),
        size = 2.5, vjust = 0, inherit.aes = FALSE
    ) +
    scale_fill_viridis_d(option = "plasma", direction = -1) +
    scale_y_continuous(expand = expansion(mult = c(0.05, 0.18))) +
    labs(
        x = "Median cluster sequence identity",
        y = "log2(Paralog SD / Allelic |diff|)",
        title = "Variance ratio by sequence identity",
        caption = "P>A = % clusters where paralog variance exceeds allelic"
    ) +
    theme(legend.position = "none", axis.text.x = element_text(angle = 45, hjust = 1))
mrv_ggsave(glue("{FIGURE_DIR}/variance_ratio_by_identity.pdf"), width = 4, height = 3.5)
```

```{r variance-ratio-by-identity-stats}
# Summary statistics table
variance_ratio_stats %>%
    mutate(
        pct_paralog_higher = round(pct_paralog_higher, 1)
    ) %>%
    knitr::kable(
        digits = 3,
        caption = "Variance ratio statistics by sequence identity",
        col.names = c(
            "Identity bin", "N clusters", "Median log2 ratio",
            "Mean log2 ratio", "% Paralog > Allelic", "y_pos"
        )
    )

# Statistical test: does variance ratio differ by identity?
cat("\n=== Kruskal-Wallis test: variance ratio ~ identity bin ===\n")
kruskal_test <- kruskal.test(
    variance_ratio_log ~ identity_bin,
    data = variance_by_identity %>% filter(is.finite(variance_ratio_log))
)
print(kruskal_test)
```

## Paralog FIRE Difference Stratified by Sequence Identity

Analyze how the difference in FIRE ratio between paralog pairs relates to their
sequence identity (gap-compressed identity from all-by-all alignments).

```{r pairwise-fire-diff}
# Calculate pairwise FIRE difference for each paralog pair
# Using the re_pairs table which has FIRE ratios for both query and target REs
re_pairs_with_fire_diff <- re_pairs %>%
    filter(
        !is.na(fire_ratio_ref_query),
        !is.na(fire_ratio_ref_target)
    ) %>%
    mutate(
        fire_diff = abs(fire_ratio_ref_query - fire_ratio_ref_target),
        # Create identity bins
        identity_bin = cut(
            gap_compressed_identity,
            breaks = c(0, 0.90, 0.95, 0.98, 0.99, 0.999, 1.0),
            labels = c("<90%", "90-95%", "95-98%", "98-99%", "99-99.9%", ">99.9%"),
            include.lowest = TRUE
        )
    )

cat("Paralog pairs with FIRE data:", nrow(re_pairs_with_fire_diff), "\n")
cat("\nPairs per identity bin:\n")
table(re_pairs_with_fire_diff$identity_bin)
```

```{r fire-diff-by-identity-violin}
# Calculate summary stats for annotations
bin_stats <- re_pairs_with_fire_diff %>%
    filter(!is.na(identity_bin)) %>%
    group_by(identity_bin) %>%
    summarise(
        n_pairs = n(),
        n_clusters = n_distinct(cluster_id),
        median_diff = median(fire_diff, na.rm = TRUE),
        y_pos = max(fire_diff, na.rm = TRUE) * 1.05,
        .groups = "drop"
    )

# Violin plot with n, cluster count, and median annotations
re_pairs_with_fire_diff %>%
    filter(!is.na(identity_bin)) %>%
    ggplot(aes(x = identity_bin, y = fire_diff, fill = identity_bin)) +
    geom_violin(alpha = 0.7, scale = "width") +
    geom_boxplot(width = 0.1, fill = "white", outlier.shape = NA) +
    geom_text(
        data = bin_stats,
        aes(
            x = identity_bin, y = y_pos,
            label = glue("n={scales::comma(n_pairs)}\nc={scales::comma(n_clusters)}\nmed={round(median_diff, 3)}")
        ),
        size = 2.5, vjust = 0, inherit.aes = FALSE
    ) +
    scale_fill_viridis_d(option = "plasma", direction = -1) +
    scale_y_continuous(expand = expansion(mult = c(0.05, 0.18))) +
    labs(
        x = "Gap-compressed sequence identity",
        y = "|FIRE ratio difference|",
        title = "Paralog FIRE divergence by sequence identity",
        caption = "n = pairs, c = clusters, med = median |diff|"
    ) +
    theme(legend.position = "none")
mrv_ggsave(glue("{FIGURE_DIR}/fire_diff_by_identity_violin.pdf"), width = 4, height = 3.5)
```

```{r fire-diff-by-identity-scatter}
# Hexbin plot of divergence (1 - identity) vs FIRE difference
# Using divergence on log10 scale spreads out high-identity pairs
# Cap at 0.0005 (99.95% identity) since we can't plot 100% on log scale
re_pairs_with_fire_diff %>%
    mutate(
        divergence = 1 - gap_compressed_identity,
        divergence = pmax(divergence, 0.0005) # Cap at 99.95% identity
    ) %>%
    ggplot(aes(x = divergence, y = fire_diff)) +
    geom_hex(bins = 50) +
    scale_fill_distiller("", palette = "Spectral", trans = "log10") +
    scale_x_continuous(
        trans = "log10",
        breaks = c(0.0005, 0.001, 0.01, 0.05, 0.1, 0.2),
        labels = c(">99.95%", "99.9%", "99%", "95%", "90%", "80%")
    ) +
    annotation_logticks(sides = "b") +
    geom_smooth(method = "lm", color = "darkred", se = FALSE, linewidth = 1) +
    stat_cor(method = "spearman", label.x.npc = "left", label.y.npc = "top", size = 3) +
    stat_regline_equation(label.x.npc = "left", label.y.npc = 0.85, size = 3) +
    labs(
        x = "Gap-compressed sequence identity (log10 divergence scale)",
        y = "|FIRE ratio difference|",
        title = "Paralog FIRE divergence vs sequence identity"
    ) +
    mrv_grid() +
    theme(legend.position = "right")
mrv_ggsave(glue("{FIGURE_DIR}/fire_diff_vs_identity_hexbin.pdf"), width = 5, height = 3)
```

```{r fire-diff-by-identity-stats}
# Summary statistics by identity bin
fire_stats_by_identity <- re_pairs_with_fire_diff %>%
    filter(!is.na(identity_bin)) %>%
    group_by(identity_bin) %>%
    summarise(
        n_pairs = n(),
        mean_fire_diff = mean(fire_diff, na.rm = TRUE),
        median_fire_diff = median(fire_diff, na.rm = TRUE),
        sd_fire_diff = sd(fire_diff, na.rm = TRUE),
        pct_diff_gt_0.1 = mean(fire_diff > 0.1, na.rm = TRUE) * 100,
        pct_diff_gt_0.2 = mean(fire_diff > 0.2, na.rm = TRUE) * 100,
        .groups = "drop"
    )

fire_stats_by_identity %>%
    knitr::kable(
        digits = 3,
        caption = "FIRE ratio difference statistics by sequence identity bin",
        col.names = c(
            "Identity", "N pairs", "Mean |diff|", "Median |diff|",
            "SD |diff|", "% diff > 0.1", "% diff > 0.2"
        )
    )
```

```{r fire-diff-correlation}
# Correlation between identity and FIRE difference
cor_test <- cor.test(
    re_pairs_with_fire_diff$gap_compressed_identity,
    re_pairs_with_fire_diff$fire_diff,
    method = "spearman"
)

cat("\n=== Correlation: Sequence Identity vs FIRE Difference ===\n")
cat("Spearman correlation:", round(cor_test$estimate, 4), "\n")
cat("p-value:", format(cor_test$p.value, scientific = TRUE, digits = 3), "\n")
```

```{r high-identity-low-fire-diff}
# What proportion of high-identity pairs (>98%) have similar FIRE ratios?
high_identity_pairs <- re_pairs_with_fire_diff %>%
    filter(gap_compressed_identity >= 0.98)

cat("\n=== High Identity Pairs (>=98% identity) ===\n")
cat("Total pairs:", nrow(high_identity_pairs), "\n")
cat("Mean FIRE diff:", round(mean(high_identity_pairs$fire_diff, na.rm = TRUE), 3), "\n")
cat("Median FIRE diff:", round(median(high_identity_pairs$fire_diff, na.rm = TRUE), 3), "\n")
cat("% with FIRE diff < 0.05:", round(mean(high_identity_pairs$fire_diff < 0.05, na.rm = TRUE) * 100, 1), "%\n")
cat("% with FIRE diff < 0.10:", round(mean(high_identity_pairs$fire_diff < 0.10, na.rm = TRUE) * 100, 1), "%\n")
cat("% with FIRE diff > 0.20:", round(mean(high_identity_pairs$fire_diff > 0.20, na.rm = TRUE) * 100, 1), "%\n")
```

```{r fire-diff-density-by-identity}
# Density plot faceted by identity bin
re_pairs_with_fire_diff %>%
    filter(!is.na(identity_bin)) %>%
    ggplot(aes(x = fire_diff, fill = identity_bin)) +
    geom_density(alpha = 0.7) +
    facet_wrap(~identity_bin, ncol = 1, scales = "free_y") +
    scale_fill_viridis_d(option = "plasma", direction = -1) +
    labs(
        x = "|FIRE ratio difference|",
        y = "Density",
        title = "Distribution of FIRE differences by identity"
    ) +
    theme(legend.position = "none")
mrv_ggsave(glue("{FIGURE_DIR}/fire_diff_density_by_identity.pdf"), width = 4, height = 5)
```

## Summary Statistics

```{r summary}
cat("=== Summary Statistics ===\n\n")
cat("Total paralog groups analyzed:", nrow(group_stats), "\n")
cat("Total paralogs in analysis:", sum(group_stats$n_paralogs), "\n\n")

cat("FIRE ratio variance (SD) across paralog groups:\n")
summary(group_stats$sd_fire_ratio)

cat("\nFIRE ratio range across paralog groups:\n")
summary(group_stats$range_fire_ratio)

cat(
    "\nProportion of groups with high variance (>2x median):",
    mean(group_stats$variance_category == "High variance"), "\n"
)
cat(
    "Proportion of groups with low variance (<0.5x median):",
    mean(group_stats$variance_category == "Low variance"), "\n"
)
```

## Session Info

```{r session-info}
sessionInfo()
```
